# CVPDC Housing Market Assessment {#part-3-1}

The following provides a regional-level analysis of major trends impacting housing within Central Virginia Planning District region. All data has been aggregated to the regional-level and includes:

- Amherst County
- Appomattox County
- Bedford County
- Campbell County
- City of Lynchburg

```{r}
#| label: setup


# Load necessary libraries.

library(tidyverse)
library(tidycensus)
library(janitor)
library(scales)
library(ggplot2)
library(tigris)
library(sf)
library(glue)
library(httr)
library(readxl)
library(hdatools)
library(ggiraph)
library(tidytext)
library(ggpattern)
library(zoo)

# Create a vector of FIPS codes of the region in order to easily pull localities from data sets.

cv <- c("51009", "51011", "51019", "51031", "51680", "51515") # CVPDC FIPS codes


```

## Takeaways



## Population trends

```{r}
#| label: pop-data
#| eval: false

# Get population estimates from the Population Estimates Program (PEP) from 2010 to 2019. 2021 data are set to be released in March 2023. 

pep_total_raw <- get_estimates( 
  geography = "county",
  state = "VA",
  variables = "POP",
  year = 2019,
  time_series = TRUE
)

# Get decennial Census estimates.

census_raw <- get_decennial( 
  geography = "county",
  state = "VA",
  year = 2020,
  sumfile = "pl",
  variables = "P1_001N"
)

pep_total_clean <- pep_total_raw %>%
  filter(!DATE %in% c(2, 3)) %>% # Remove non-decennial 2010 counts
  mutate(year = # Translate date codes into years
    case_when(
      DATE == 1 ~ "2010",
      DATE == 4 ~ "2011",
      DATE == 5 ~ "2012",
      DATE == 6 ~ "2013",
      DATE == 7 ~ "2014",
      DATE == 8 ~ "2015",
      DATE == 9 ~ "2016",
      DATE == 10 ~ "2017",
      DATE == 11 ~ "2018",
      DATE == 12 ~ "2019")) %>%
  mutate(counttype = # Add descriptions to count types
      case_when(
        DATE == 1 ~ "Census population",
        TRUE ~ "Population estimate")) %>% 
  select( # Simplify columns
    GEOID,
    year,
    counttype,
    value
  )

# Prep total population counts from 2020 Census summary file

census_clean <- census_raw %>% 
  mutate(year = "2020", # Add year and count type columns
         counttype = "Census population") %>% 
  select( # Simplify columns
    GEOID,
    year,
    counttype,
    value
  )

population_data <-pep_total_clean %>% 
  bind_rows(census_clean) %>% 
  subset(GEOID %in% cv)

write_rds(population_data, "data/pop_data.rds")

```

```{r}
#| label: fig-pop
#| fig-cap: "Total regional population"

# Read in saved data and aggregate to the regional level.

cv_pop <- read_rds("data/pop_data.rds") %>% 
  group_by(year, counttype) %>% # Group by year and the count type.
  summarise(value = sum(value)) # Sum the values.

# Create bar chart showing the change in population from 2010 to 2020

pop_bar <- ggplot(cv_pop, # Read in the data.
                  aes(x = year, # Plot the x-axis.
                      y = value, # Plot the y-axis.
                      fill = value, # Set the variable that will be colored.
                      data_id = value, # Set the variable that will be used in interactive plot.
                      tooltip = label_comma()(value))) + # Set the format of the data label on hover.
  geom_col() + # Set as a bar chart. 
  geom_col_interactive(size = 2) +
  theme_hfv() + # Apply custom HFV theme.
  scale_fill_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) + # Use custom HFV gradient.
  labs(title = "Total population",
       subtitle = "From 2010 to 2020",
       caption = "**Source:** U.S. Census Bureau, Population Estimates Program (PEP).") +
  scale_y_continuous(label = label_comma()) # Format the y-axis to show commas.

# Apply if-else statement to print interactive plot if in html or simply print the visualization static plot if anything else.
  
if (knitr::is_html_output()) {
  
  girafe(ggobj = pop_bar,
         height_svg = 4)
} else {
  
  pop_bar
}
  

```

```{r}
#| label: components-data
#| eval: false

# PEP does not have data available for 2020 and 2021. There is a gap between 2019 and 2021 because Census uses 2020 as a baseline. 

pep_change_raw <- get_estimates(
  geography = "county",
  state = "VA",
  variables = c("NATURALINC", "DOMESTICMIG", "INTERNATIONALMIG"),
  time_series = TRUE
)

pep_change_clean <- pep_change_raw %>%  
  mutate(GEOID = as.character(GEOID)) %>%  
  filter(GEOID %in% cv) %>% 
  mutate(year = # Translate date codes into years
    case_when(
      PERIOD == 1 ~ "2010",
      PERIOD == 2 ~ "2011",
      PERIOD == 3 ~ "2012",
      PERIOD == 4 ~ "2013",
      PERIOD == 5 ~ "2014",
      PERIOD == 6 ~ "2015",
      PERIOD == 7 ~ "2016",
      PERIOD == 8 ~ "2017",
      PERIOD == 9 ~ "2018",
      PERIOD == 10 ~ "2019")) %>% 
  mutate(component = # Rename components of change
    case_when(
      variable == "NATURALINC" ~ "Natural increase",
      variable == "DOMESTICMIG" ~ "Domestic migration",
      variable == "INTERNATIONALMIG" ~ "International migration")) %>%  
  select( # Simplify columns
    GEOID,
    year,
    component,
    value
  )

pep_2021 <- read_csv("data/comp_chang_2021.csv") %>% 
  pivot_longer(4:6,
               names_to = "component",
               values_to = "value") %>% 
  mutate(year = 2021,
         component =
    case_when(
      component == "NATURALINC" ~ "Natural increase",
      component == "DOMESTICMIG" ~ "Domestic migration",
      component == "INTERNATIONALMIG" ~ "International migration")) %>% 
  select(GEOID, year, component, value) |> 
  mutate(GEOID = as.character(GEOID))

cv_comp_change <- pep_change_clean %>% 
  rbind(pep_2021) |>
  filter(GEOID %in% cv) %>%
  left_join(counties("VA", year = 2021), by = "GEOID") %>%
  select(NAMELSAD, year, component, value)

write_rds(cv_comp_change, "data/comp_data.rds")

```

```{r}

# Read in saved data. 

cv_comp_change <- read_rds("data/comp_data.rds")

# Aggregate data to the regional level.

cv_change <- cv_comp_change %>% 
  group_by(year, component) %>% # Group by year and component of change.
  summarise(value = sum(value)) # Sum values.

cv_change_plot <- ggplot(cv_change,
       aes(x = year,
           y = value,
           fill = component,
           data_id = value,
           tooltip = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")(value))) +
  geom_col(position = "stack") +
  geom_col_interactive(size = 2) +
  theme_hfv() +
  scale_fill_hfv() + 
  labs(title = "Components of population change",
       subtitle = "From 2010 to 2021",
       caption = "**Source:** U.S. Census Bureau, Population Estimates Program (PEP).") +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus"))
  
if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_change_plot,
         height_svg = 4)
} else {
  
  cv_change_plot
}

```

```{r}
#| label: fig-projections

# Read in saved UVA Projections data.

projections <- read_csv("data/uva_proj.csv") %>% 
  subset(FIPS %in% cv) %>% # Filter for Lynchburg metro localities
  select(FIPS, NAME, `2030`,`2040`, `2050`) %>%  # Select only necessary variables
  pivot_longer(3:5, # Choose which columns to pivot
               names_to = "year", # Name the variable
               values_to = "projection") # Name the data value


dec_2020 <- c("2020", 261593) # 2020 Decennial Census for context.

# Aggregate the data to the region. 

cv_proj <- projections %>% 
  group_by(year) %>% 
  summarise(count = sum(projection)) %>% 
  rbind(dec_2020) %>% # Add 2020 Decennial Census data to data frame.
  arrange(year) # Arrange in ascending year order to visualize properly. 

cv_proj$type <- c("Census count", "Projection", "Projection", "Projection") # Add column to describe count type.

cv_proj <- cv_proj %>% 
  transform(count = as.numeric(count)) 

cv_proj_plot <- ggplot(cv_proj,
       aes(x = year,
           y = count,
           fill = type)) +
  geom_col() + 
  theme_hfv() + 
  scale_fill_hfv() +
  geom_text(aes(label = scales::comma(count)), 
            size = 4, 
            vjust = 1.5, 
            color = "white", 
            fontface = "bold") +
  labs(title = "Population projection",
       caption = "**Source:** University of Virginia, Weldon Cooper Center.") +
  scale_y_continuous(labels = label_comma())


cv_proj_plot


```

## Household trends

```{r}
#| label: tenure-data
#| eval: false

# Table B25003: Households by Tenure

years <- 2010:2021

b25003_vars <- load_variables(2021, "acs5") %>% 
  filter(str_sub(name, end = 6) %in% "B25003") %>% 
  filter(str_length(name) < 11)

b25003_raw <- map_dfr(years, function(yr){
  b25003_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B25003",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>% 
    mutate(year = yr)
})

b25003_raw <- b25003_raw %>% 
  subset(GEOID %in% cv)


b25003_vars_cleaned <- b25003_vars %>% 
  separate(label, into = c("est", "total", "tenure"), sep = "!!") %>% 
  select(variable = name, tenure) |>
  drop_na() |>
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  ))

b25003_data <- b25003_raw %>% 
  right_join(b25003_vars_cleaned, by = "variable") %>% 
  select(NAME, GEOID, year, tenure, estimate, moe) %>% 
  mutate(NAME = str_remove_all(NAME, ", Virginia"),
         NAME = case_when(
           NAME == "Bedford city" ~ "Bedford County", # Aggregate Bedford City and Bedford County
           TRUE ~ NAME
         ))

write_rds(b25003_data, "data/b25003_data.rds")

```

```{r}
#| label: fig-tenure

# Read in saved data. 

b25003_data <- read_rds("data/b25003_data.rds") 

# Aggregate data and conduct analysis to identify the cumulative change.

cv_tenure <- b25003_data %>% 
  group_by(year, tenure) %>% # Group by year and tenure to aggregate to region.
  summarise(estimate = sum(estimate)) %>% # Summarize data for the aggregation.
  group_by(tenure) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) %>% 
  mutate(year = as.character(year))

cv_tenure_plot <- ggplot(cv_tenure,
       aes(x = year, 
           y = run_diff,
           fill = tenure,
           data_id = run_diff,
           tooltip = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")(run_diff))) +
  geom_col() +
  geom_col_interactive(size = 2) +
  facet_wrap(~tenure) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")) +
  theme(axis.text.x = element_text(angle = 90, vjust = .2)) +
  labs(title = "Cumulative change in households by tenure", 
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25003.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_tenure_plot,
         height_svg = 4)
} else {
  
  cv_tenure_plot
}

```

```{r}
#| label: age
#| eval: false

# Table B25007: Tenure by Age of Householder

b25007_vars <- load_variables(2021, "acs5") %>%
  filter(str_sub(name, end = 6) %in% "B25007")

b25007_raw <- map_dfr(years, function(yr){
  b25007_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B25007",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
})

b25007_raw <- b25007_raw %>%
  subset(GEOID %in% cv)

b25007_vars_cleaned <- b25007_vars %>%
  separate(label, into = c("est", "total", "tenure", "age"), sep = "!!") %>%
  select(variable = name, tenure, age) %>%
  drop_na() |>
  mutate(tenure = case_when(
    tenure == "Owner occupied:" ~ "Homeowner",
    tenure == "Renter occupied:" ~ "Renter"
  ))

b25007_data <- b25007_raw %>%
  right_join(b25007_vars_cleaned, by = "variable") %>% 
  select(NAME, GEOID, year, tenure, age, estimate, moe) %>% 
  mutate(NAME = str_remove_all(NAME, ", Virginia"),
         NAME = case_when(
           NAME == "Bedford city" ~ "Bedford County", # Aggregate Bedford City and Bedford County
           TRUE ~ NAME),
         age = case_when(
    age == "Householder 15 to 24 years" ~ "24 years and under",
    age == "Householder 25 to 34 years" ~ "25 to 44 years old",
    age == "Householder 35 to 44 years" ~ "25 to 44 years old",
    age == "Householder 45 to 54 years" ~ "45 to 64 years old",
    age == "Householder 55 to 59 years" ~ "45 to 64 years old",
    age == "Householder 60 to 64 years" ~ "45 to 64 years old",
    age == "Householder 65 to 74 years" ~ "65 years and over",
    age == "Householder 75 to 84 years" ~ "65 years and over",
    age == "Householder 85 years and over" ~ "65 years and over"
  ))

write_rds(b25007_data, "data/b25007_data.rds")

```

```{r}
#| label: fig-age

b25007_data <- read_rds("data/b25007_data.rds")

cv_age <- b25007_data %>% 
  group_by(NAME, year, tenure, age) %>% 
  summarise(estimate = sum(estimate)) %>% 
  pivot_wider(names_from = year,
              values_from = estimate) %>% 
  select(tenure, age, `2010`, `2021`) %>% 
  transform(change = `2021` - `2010`) %>% 
  group_by(NAME, tenure, age) %>% 
  summarise(change = sum(change)) %>% 
  mutate(age = fct_relevel(age, "Under 25 years old")) %>% 
  group_by(tenure, age) %>% 
  summarise(change = sum(change))

cv_age_plot <- ggplot(cv_age,
       aes(x = tenure, 
           y = change,
           fill = tenure,
           data_id = change,
           tooltip = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")(change)))  +
  facet_wrap(~age, nrow =1) +
  geom_col(position = "dodge") +
  geom_col_interactive(size = 2) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")) +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
  labs(title = "Change in households by age and tenure",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25007.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_age_plot,
         height_svg = 4)
} else {
  
  cv_age_plot
}


```


```{r}
#| label: hhtype-data
#| eval: false

# Table B11001: Household type (including Living Alone)

b11001_vars <- load_variables(2021, "acs5") %>% 
  filter(str_sub(name, end = 7) %in% "B11001_")


b11001_raw <- map_dfr(years, function(yr){
  b11001_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B11001",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |>
    mutate(year = yr)
})

b11001_raw <- b11001_raw %>% 
  subset(GEOID %in% cv)

b11001_vars_cleaned <- b11001_vars %>% 
  separate(label, into = c("est", "tot", "type", "relationship", "householder"),
           sep = "!!") |>
  select(variable = name, type, relationship, householder) %>% 
  mutate(
    householder = case_when(
      relationship == "Married-couple family" ~ relationship,
      relationship == "Householder living alone" ~ relationship,
      relationship == "Householder not living alone" ~ relationship,
      TRUE ~ householder),
    relationship = case_when(
      relationship == "Householder living alone" ~ type,
      relationship == "Householder not living alone" ~ type,
      TRUE ~ relationship)
    ) %>% 
  mutate(across(.fns = ~str_remove_all(.x, ":"))) %>% 
  drop_na()

b11001_data <- b11001_raw |>
  right_join(b11001_vars_cleaned, by = "variable") |>
  select(NAME, GEOID, year, "hhtype" = householder, relationship, type, estimate) |>
  mutate(NAME = str_remove_all(NAME, ", Virginia"),
          NAME = case_when(
           NAME == "Bedford city" ~ "Bedford County", # Aggregate Bedford City and Bedford County
           TRUE ~ NAME))

write_rds(b11001_data, "data/b11001_data.rds")

```

```{r}
#| label: fig-hhtype

b11001_data <- read_rds("data/b11001_data.rds")

cv_hhtype <- b11001_data %>% 
  group_by(year, hhtype) %>% 
  summarise(estimate = sum(estimate)) %>% 
  pivot_wider(names_from = year,
              values_from = estimate) %>% 
  select(hhtype, `2010`, `2021`) %>% 
  transform(change = `2021` - `2010`) 


cv_hhtype_plot <-ggplot(cv_hhtype,
       aes(x = reorder(hhtype, change),
           y = change,
           fill = hhtype)) +
  geom_col() +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_comma()) + 
   scale_x_discrete(labels = function(x) str_wrap(x, width = 16)) +
  labs(title = "Change in household type",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B11001.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_tenure_plot,
         height_svg = 4)
} else {
  
  cv_tenure_plot
}

```


```{r}
#| label: hhsize-data
#| eval: false

# Table B25009: Household size by tenure

b25009_vars <- load_variables(2021, "acs5") %>% 
  filter(str_sub(name, end = 6) %in% "B25009")

b25009_raw <- map_dfr(years, function(yr){
  b25009_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B25009",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |>
    mutate(year = yr)
})

b25009_raw <- b25009_raw %>% 
  subset(GEOID %in% cv)

b25009_vars_cleaned <- b25009_vars %>% 
  separate(label, into = c("est", "total", "tenure", "size"), sep = "!!") %>% 
  select(variable = name, tenure, size)  %>% 
  drop_na() |>
  mutate(tenure = case_when(
    tenure == "Owner occupied:" ~ "Homeowner",
    tenure == "Renter occupied:" ~ "Renter"
  ))

b25009_data <- b25009_raw %>% 
  right_join(b25009_vars_cleaned, by = "variable") %>% 
  select(NAME, GEOID, year, tenure, size, estimate) %>% 
  mutate(NAME = str_remove_all(NAME, ", Virginia"),
         NAME = case_when(
           NAME == "Bedford city" ~ "Bedford County", # Aggregate Bedford City and Bedford County
           TRUE ~ NAME)) %>% 
  mutate(size = case_when(
    size == "1-person household" ~ "1-person",
    size == "2-person household" ~ "2-person",
    size == "3-person household" ~ "3-person",
    TRUE ~ "4 or more persons"
  )) 

write_rds(b25009_data, "data/b25009_data.rds")

```

```{r}
#| label: fig-hhsize

b25009_data <- read_rds("data/b25009_data.rds")

cv_hhsize <- b25009_data %>% 
  group_by(year, tenure, size) %>% 
  summarise(estimate = sum(estimate)) %>% 
  pivot_wider(names_from = year,
              values_from = estimate) %>% 
  select(tenure, size, `2010`, `2021`) %>% 
  transform(change = `2021` - `2010`) 
  

cv_hhsize_plot <- ggplot(cv_hhsize,
       aes(x = tenure,
           y = change,
           fill = tenure,
           data_id = change,
           tooltip = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")(change))) +
  geom_col() +
  geom_col_interactive(size = 2)
  facet_wrap(~size, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Change in household size by tenure",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25009.") +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus"))
  
if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_hhsize_plot,
         height_svg = 4)
} else {
  
  cv_hhsize_plot
}

```

```{r}
#| label: hh-seniors-data
#| eval: false

# Table B09020: Senior population by living arrangements

years <- 2012:2021 # 5-year estimates only go back to 2012

b09020_vars <- load_variables(2021, "acs5") |>
  filter(str_sub(name, end = 6) %in% "B09020")

b09020_raw <- map_dfr(years, function(yr){
  b09020_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B09020",
    year = yr,
    survey = "acs5"
  ) |>
    mutate(year = yr)
})

b09020_raw <- b09020_raw |>
  subset(GEOID %in% cv)

b09020_vars_cleaned <- b09020_vars |>
  separate(label, into = c("est", "total", "hhgq", "family", "relationship", "sex", "alone"), sep = "!!") |>
  select(variable = name, hhgq, family, relationship, alone) |>
  filter(!variable %in% c("B09020_001", "B09020_002", "B09020_003", "B09020_005", "B09020_006",
                          "B09020_012", "B09020_013", "B09020_014", "B09020_017")) |>
  mutate(relationship = case_when(
    relationship == "Spouse" ~ "With spouse",
    relationship == "Nonrelatives" ~ "With nonrelatives",
    relationship %in% c("Parent", "Parent-in-law", "Other relatives") ~ "With other relative(s)",
    hhgq == "In group quarters" ~ "Group quarters",
    !is.na(alone) ~ alone,
    TRUE ~ relationship
  )) |>
  mutate(family = case_when(
    family == "In family households:" ~ "Family",
    family == "In nonfamily households:" ~ "Nonfamily",
    hhgq == "In group quarters" ~ "Group quarters"
  )) |>
  mutate(across(.fns = ~str_remove_all(.x, ":"))) |>
  select(1,3,4)

b09020_data <- b09020_raw |>
  right_join(b09020_vars_cleaned, by = "variable") |>
  select(NAME, GEOID, year, family, relationship, estimate) |>
  mutate(NAME = str_remove_all(NAME, ", Virginia"),
         NAME = case_when(
           NAME == "Bedford city" ~ "Bedford County", # Aggregate Bedford City and Bedford County
           TRUE ~ NAME))
  
write_rds(b09020_data, "data/b09020_data.rds")

```

```{r}
#| label: fig-seniors

b09020_data <- read_rds("data/b09020_data.rds")

cv_seniors <- b09020_data %>% 
  group_by(year, family, relationship) %>% 
  summarise(estimate = sum(estimate)) %>% 
  pivot_wider(names_from = year,
              values_from = estimate) %>% 
  transform(change = `2021` - `2012`) %>% 
  mutate(family = str_replace(family, "Group quarters", " "),
         family = fct_relevel(family, "Family", "Nonfamily", " "))


cv_seniors_plot <- ggplot(cv_seniors,
       aes(y = reorder_within(relationship, change, family),
           x = change,
           fill = family,
           data_id = change,
           tooltip = label_comma()(change))) +
  geom_col(position = "dodge") +
  facet_grid(rows = vars(family), scales = "free_y", space = "free", switch = "y") +
  scale_y_reordered() +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Change in senior population by living arrangement",
       caption = "**Source:** U.S. Census Bureau, American Community Survey, Table B09020.") +
  theme(strip.placement = "outside",
        strip.text.y = element_text(angle = 0),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(color = "#e2e4e3",
                                          size = 0.05),
        plot.title.position = "plot",
        plot.caption.position = "plot"
        ) +
   scale_x_continuous(labels = label_comma())

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_seniors_plot,
         height_svg = 4)
} else {
  
  cv_seniors_plot
}
```


```{r}
#| label: hh-subfam-data
#| eval: false

# Table B11013: Subfamily type by presence of own children

years <- 2010:2021

b11013_vars <- load_variables(2021, "acs5") |>
  filter(str_sub(name, end = 6) %in% "B11013")

b11013_raw <- map_dfr(years, function(yr){
  b11013_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B11013",
    year = yr,
    survey = "acs5"
  ) |>
    mutate(year = yr)
})

b11013_raw <- b11013_raw |>
  subset(GEOID %in% cv)

b11013_vars_cleaned <- b11013_vars |>
  separate(label, into = c("est", "total", "subfamily", "children"), sep = "!!") |>
  select(variable = name, subfamily, children) |>
  filter(!variable %in% c("B11013_001", "B11013_002")) |>
  mutate(subfamily = case_when(
    subfamily == "Married-couple subfamily:" ~ "Married couple",
    subfamily == "Mother-child subfamily" ~ "Single mother",
    subfamily == "Father-child subfamily" ~ "Single father")) |>
  mutate(children = case_when(
    children == "With own children under 18 years" ~ "With children",
    children == "No own children under 18 years" ~ "No children",
    TRUE ~ "With children"))

b11013_data <- b11013_raw |>
  right_join(b11013_vars_cleaned, by = "variable") |>
  select(NAME, GEOID, year, subfamily, children, estimate) |>
  mutate(NAME = str_remove_all(NAME, ", Virginia"),
         NAME = case_when(
           NAME == "Bedford city" ~ "Bedford County", # Aggregate Bedford City and Bedford County
           TRUE ~ NAME))

write_rds(b11013_data, "data/b11013_data.rds")

```


```{r}
#| label: fig-subfam

b11013_data <- read_rds("data/b11013_data.rds") 

cv_subfam <- b11013_data %>% 
  group_by(year, subfamily, children) %>%  
  summarise(estimate = sum(estimate)) %>% 
  mutate(year = as.character(year))

cv_subfam_plot <- ggplot(cv_subfam,
       aes(x = year,
           y = estimate,
           fill = children,
           data_id = estimate,
           tooltip = label_comma()(estimate))) +
  geom_col(position = "dodge") +
  facet_wrap(~subfamily) +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Subfamilies by type and presence of own children",
       caption = "**Source:** U.S. Census Bureau, American Community Survey, Table B11013.") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = .2))

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_subfam_plot,
         height_svg = 4)
} else {
  
  cv_subfam_plot
}
  

```


```{r} 
#| label: hh-multigen-data
#| eval: false

library(sf)
# Load PUMA geographies for Virginia
pumas <- pumas(state = "VA")

# Filter PUMAs to Virginia Region 2000 (North and South)

pumas_cv <- pumas |>
  filter(str_detect(NAMELSAD10, "Lynchburg|Bedford"))

# Pull each PUMS year separately because data structure is not same across 2016 to 2020

multigen_2016 <- get_pums(
  variables = "MULTG",
  state = "VA",
  year = 2016,
  puma = pumas_cv$PUMACE10)

multigen_2016 <- multigen_2016 |>
  mutate(YEAR = "2016",
         SPORDER = as.numeric(SPORDER),
         PUMA = case_when(
           str_length(PUMA) == 4 ~ paste0("0", PUMA),
           TRUE ~ as.character(PUMA))) |>
  select(YEAR, SERIALNO, PUMA, SPORDER, WGTP, PWGTP, MULTG)

multigen_2017 <- get_pums(
  variables = "MULTG",
  state = "VA",
  year = 2017,
  puma = pumas_cv$PUMACE10)

multigen_2017 <- multigen_2017 |>
  mutate(SPORDER = as.numeric(SPORDER),
         YEAR = "2017") |>
  select(YEAR, SERIALNO, PUMA, SPORDER, WGTP, PWGTP, MULTG)

multigen_2018 <- get_pums(
  variables = "MULTG",
  state = "VA",
  year = 2018,
  puma = pumas_cv$PUMACE10)

multigen_2018 <- multigen_2018 |>
  mutate(SPORDER = as.numeric(SPORDER),
         YEAR = "2018") |>
  select(YEAR, SERIALNO, PUMA, SPORDER, WGTP, PWGTP, MULTG)

multigen_2019 <- get_pums(
  variables = "MULTG",
  state = "VA",
  year = 2019,
  puma = pumas_cv$PUMACE10)

multigen_2019 <- multigen_2019 |>
  mutate(YEAR = "2019") |>
  select(YEAR, SERIALNO, PUMA, SPORDER, WGTP, PWGTP, MULTG)

multigen_2020 <- get_pums(
  variables = "MULTG",
  state = "VA",
  year = 2020,
  puma = pumas_cv$PUMACE10)

multigen_2020 <- multigen_2020 |>
  mutate(YEAR = "2020") |>
  select(YEAR, SERIALNO, PUMA, SPORDER, WGTP, PWGTP, MULTG)

# Bind years together and recode

multigen_raw <- bind_rows(multigen_2016, multigen_2017, multigen_2018,
                          multigen_2019, multigen_2020) |>
  filter(!MULTG %in% c("b", "0")) |>
  mutate(geography = case_when(
    PUMA == "51096" ~ "Lynchburg city and Campbell County",
    PUMA == "51095" ~ "Amherst, Appomattox, and Bedford Counties"
  )) %>% 
  mutate(MULTG = case_when(
    MULTG == "1" ~ "Single generation",
    MULTG == "2" ~ "Multiple generations"
  ))

# Summarize percent of multigenerational households by year
multigen_data <- multigen_raw |>
  group_by(geography, YEAR) |>
  summarise(pct = sum(PWGTP[MULTG == "Multiple generations"]) / sum(PWGTP))|>
  mutate(geography = fct_reorder(geography, pct, .desc = TRUE))

write_rds(multigen_data, "data/multigen_data.rds")

```

```{r}
#| label: fig-multigen

multigen_data <- read_rds("data/multigen_data.rds")


multigen_plot <- ggplot(multigen_data,
       aes(x = YEAR,
           y = pct,
           color = geography,
           group = 1)) +
  geom_line(size = 1) +
  geom_point_interactive(
    aes(data_id = pct,
        tooltip = label_percent(accuracy = 0.1)(pct)),
    size = 1) +
  facet_wrap(~geography, nrow = 1) +
  scale_y_continuous(labels = label_percent(),
                     breaks = c((1:9)/100),
                     limits = c(0.005,0.09)) +
  labs(title = "Percent of population living in multiple generation households",
       subtitle = "2016 to 2020",
       caption = "**Source:** U.S. Census Bureau, American Community Survey, Public Use Microdata Sample.") +
  scale_color_pha() +
  theme_pha() +
  theme(axis.text.x = element_text(angle = 90))

if (knitr::is_html_output()) {
  
  girafe(ggobj = multigen_plot,
         height_svg = 4)
} else {
  
  multigen_plot
}


```

```{r}
#| label: hh-adultchild-data
#| eval: false

years <- 2015:2021

b09021_vars <- load_variables(2021, "acs5") |>
  filter(str_sub(name, end = 6) %in% "B09021")

b09021_raw <- map_dfr(years, function(yr){
  b09021_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B09021",
    year = yr,
    survey = "acs5"
  ) |>
    mutate(year = yr)
})

b09021_raw <- b09021_raw |>
  subset(GEOID %in% cv)

b09021_vars_cleaned <- b09021_vars |>
  filter(name %in% c("B09021_009", "B09021_010", "B09021_011",
                     "B09021_012", "B09021_013", "B09021_014")) |>
  separate(label, into = c("est", "total", "age", "arrangement"), sep = "!!") |>
  select(variable = name, arrangement) |>
  mutate(arrangement = case_when(
    arrangement == "Householder living with spouse or spouse of householder" ~ "With spouse",
    arrangement == "Householder living with unmarried partner or unmarried partner of householder" ~ "With partner",
    TRUE ~ arrangement
  ))

b09021_data <- b09021_raw |>
  right_join(b09021_vars_cleaned, by = "variable") |>
  select(NAME, GEOID, year, arrangement, estimate) |>
  mutate(NAME = str_remove_all(NAME, ", Virginia"),
         NAME = case_when(
           NAME == "Bedford city" ~ "Bedford County", # Aggregate Bedford City and Bedford County
           TRUE ~ NAME)) 


write_rds(b09021_data, "data/b09021_data.rds")
```


```{r}
#| label: fig-adultchild

b09021_data <- read_rds("data/b09021_data.rds") 

cv_adultchild <- b09021_data %>% 
  group_by(year, arrangement) %>% 
  summarise(estimate = sum(estimate)) %>% 
  filter(year %in% c(2015, 2021)) %>% 
  pivot_wider(names_from = year,
              values_from = estimate) %>% 
  mutate(change = `2021` - `2015`,
         pct_change = change/`2015`)

cv_adultchild_plot <- ggplot(cv_adultchild,
       aes(x = reorder(arrangement, pct_change),
           y = pct_change,
           fill = arrangement,
           label = label_percent()(pct_change),
           data_id = pct_change,
           tooltip = label_percent(accuracy = 0.1)(pct_change))) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge", size = 2) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(title = "Change in 18-34 year old population by living arrangement",
       subtitle="2015 to 2021",
       caption = "**Source:** U.S. Census Bureau, American Community Survey, Table B09021.") +
  scale_y_continuous(labels = percent_format())

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_adultchild_plot,
         height_svg = 4)
} else {
  
  cv_adultchild_plot
}

```


## Economic trends

```{r}
#| label: hh-inc-data
#| eval: false

years <- 2010:2021

b25118_vars <- load_variables(2021, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25118")

b25118_raw <- map_dfr(years, function(yr){
  b25118_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B25118",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |> 
    mutate(year = yr)
  })

b25118_raw <- b25118_raw |> 
  subset(GEOID %in% cv)

b25118_vars_cleaned <- b25118_vars |> 
  separate(label, into = c("est", "total", "tenure", "income"), sep = "!!") |>  
  select(variable = name, tenure, income) |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied:" ~ "Homeowner",
    tenure == "Renter occupied:" ~ "Renter"
  )) |> 
  drop_na()

b25118_data <- b25118_raw |> 
  right_join(b25118_vars_cleaned, by = "variable") |> 
  select(NAME, year, tenure, income, estimate) |> 
  mutate(NAME = str_remove_all(NAME, ", Virginia"),
         NAME = case_when(
           NAME == "Bedford city" ~ "Bedford County", # Aggregate Bedford City and Bedford County
           TRUE ~ NAME)) |>
  mutate(income = case_when(
    income == "Less than $5,000" ~ "Less than $15,000",
    income == "$5,000 to $9,999" ~ "Less than $15,000",
    income == "$10,000 to $14,999" ~ "Less than $15,000",
    income == "$15,000 to $19,999" ~ "$15,000 to $24,999",
    income == "$20,000 to $24,999" ~ "$15,000 to $24,999",
    income == "$25,000 to $34,999" ~ "$25,000 to $49,999",
    income == "$35,000 to $49,999" ~ "$25,000 to $49,999",
    income == "$50,000 to $74,999" ~ "$50,000 to $74,999",
    TRUE ~ income
  ))
  
write_rds(b25118_data, "data/b25118_data.rds")
```

```{r}

b25118_data <- read_rds("data/b25118_data.rds") 

inc_dist <- b25118_data %>% 
  group_by(year, tenure, income) %>% 
  summarise(estimate = sum(estimate)) %>% 
  filter(year == 2010 | year == 2021) %>% 
  mutate(year = as.character(year)) 

level_order <- c("Less than $15,000", "$15,000 to $24,999", "$25,000 to $49,999", "$50,000 to $74,999", "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more")

inc_dist_plot <- ggplot(inc_dist,
       aes(x = factor(income, level = level_order),
           y = estimate,
           fill = tenure,
           data_id = estimate,
           tooltip = label_comma()(estimate))) +
  geom_col(position = "dodge") +
  geom_col_interactive(size = 1) +
  facet_grid(tenure~year) +
  scale_fill_hfv() +
  theme_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = .2, size = 8),
        strip.text.y = element_blank(),
        legend.position = "bottom") +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Income distribution",
       subtitle = "In 2010 and 2021") 

if (knitr::is_html_output()) {
  
  girafe(ggobj = inc_dist_plot,
         height_svg = 4)
} else {
  
  inc_dist_plot
}

```

```{r}
#| label: med-income-msa
#| eval: false


# "metropolitan statistical area/micropolitan statistical area"

years <- 2010:2021 

b25119_vars <- load_variables(2021, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25119")


b25119_raw <- map_dfr(years, function(yr){
  b25119_pull <- get_acs(
    geography = "cbsa",
    table = "B25119",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })  


b25119_msa <- b25119_raw %>% 
  filter(NAME == "Lynchburg, VA Metro Area")

b25119_vars_cleaned <- b25119_vars %>% 
  separate(label, into = c("est", "income", "total", "tenure"), sep = "!!") %>% 
  select(variable = name, tenure) %>% 
  mutate(tenure = case_when(
    tenure == "Owner occupied (dollars)" ~ "Homeowner",
    tenure == "Renter occupied (dollars)" ~ "Renter"
  ))

b25119_vars_cleaned$tenure <- b25119_vars_cleaned$tenure %>%  
  replace_na('All households')

b25119_data <- b25119_msa %>% 
  right_join(b25119_vars_cleaned, by = "variable") |> 
  select(NAME, year, tenure, estimate, moe)

library(readxl)

cpi <- read_excel("data/CPI_U_RS.xlsx")

cpi <- cpi %>% 
  rename(year = Year,
         priceindex = Index) |> 
  transform(year = as.numeric(year))

b25119_cpi_msa <- b25119_data |> 
  left_join(cpi, by = 'year') |> 
  transform(dollars21 = ((399/priceindex)*estimate)) |> 
  select(NAME, year, tenure, dollars21, cdollars = estimate) |>
  filter(tenure != "All households") |> 
  select(year, NAME, tenure, dollars21, cdollars)

write_rds(b25119_cpi_msa, "data/b25119_cpi_msa.rds")

```

```{r}

b25119_cpi_msa <- read_rds("data/b25119_cpi_msa.rds") %>% 
  pivot_longer(4:5,
               names_to = "value",
               values_to = "estimate") %>% 
  mutate(value = case_when(
    value == "dollars21" ~ "2021 inflation-adjusted",
    value == "cdollars" ~ "Current dollars"
  ))

med_inc_plot <- ggplot(b25119_cpi_msa,
       aes(x = year,
           y = estimate,
           color = tenure,
           linetype = value)) +
  geom_line(size = .5) +
  geom_point_interactive(
    aes(data_id = estimate,
        tooltip = dollar_format()(estimate)),
    size = 1) +
  facet_wrap(~tenure) +
  theme_hfv() +
  scale_color_hfv() +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Median household income by tenure",
       subtitle = "From 2010 to 2021",
       caption = "**Source:** U.S. Census Bureau, American Community Survey, 5-Year Estimates, Table B25119.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = med_inc_plot,
         height_svg = 4)
} else {
  
  med_inc_plot
}

```


```{r}
#| label: med-income-county
#| eval: false


years <- 2010:2021 #Including Bedford City, but be mindful that you cannot aggregate medians.

b25119_vars <- load_variables(2021, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25119")

b25119_raw <- map_dfr(years, function(yr){
  b25119_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B25119",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25119_raw <- b25119_raw |> 
  subset(GEOID %in% cv)

b25119_vars_cleaned <- b25119_vars |> 
  separate(label, into = c("est", "income", "total", "tenure"), sep = "!!") |>  
  select(variable = name, tenure) |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied (dollars)" ~ "Homeowner",
    tenure == "Renter occupied (dollars)" ~ "Renter"
  ))

b25119_vars_cleaned$tenure <- b25119_vars_cleaned$tenure |>  replace_na('All households')
b25119_data <- b25119_raw |> 
  right_join(b25119_vars_cleaned, by = "variable") |> 
  select(NAME, GEOID, year, tenure, estimate, moe)


cpi <- read_excel("data/CPI_U_RS.xlsx")

cpi <- cpi |> 
  rename(year = Year,
         priceindex = Index) |> 
  transform(year = as.numeric(year))

b25119_cpi <- b25119_data |> 
  left_join(cpi, by = 'year') |> 
  transform(dollars21 = ((399/priceindex)*estimate)) |> 
  select(NAME, GEOID, year, tenure, dollars21, cdollars = estimate) |>
  filter(tenure != "All households") |> 
  select(year, NAME, tenure, dollars21, cdollars)

write_rds(b25119_cpi, "data/b25119_cpi.rds")

```

```{r}
#| label: inc-race-msa
#| eval: false

years <- 2010:2021 

b19013 <- paste0("B19013", LETTERS[2:9])

b19013_defns <- load_variables(2020, "acs5") %>%
  filter(str_sub(name, end = 7) %in% b19013) %>%
  filter(str_detect(name, "PR") == FALSE)

concept_to_race <- function(x) {
  out <- x %>%
    str_remove_all("MEDIAN HOUSEHOLD INCOME IN THE PAST 12 MONTHS \\(\\IN 2020 INFLATION-ADJUSTED DOLLARS\\)\\ \\(|\\)") %>%
    str_to_title()
}

b19013_cleaned <- b19013_defns %>%
  mutate(race = concept_to_race(concept)) %>%
  separate(label, c("estimate", "medhhincome"), sep = "!!") %>%
  select(variable = name, medhhincome, race) %>%
  mutate(across(.fns = ~replace_na(.x, "All")),
          across(.fns = ~str_remove_all(.x, ":")))

b19013_raw_msa <- map_dfr(b19013, function(tb) {
  yearly_data <- map_dfr(years, function(yr) {
    acs_pull <- get_acs(
      geography = "cbsa",
      table = tb,
      year = yr
    ) %>%
    left_join(b19013_cleaned, by = "variable")
    acs_rearranged <- acs_pull %>%
      mutate(year = yr) %>%
      select(variable, year, NAME, GEOID, race, medhhincome,
             estimate, moe)
    acs_rearranged
  })
  yearly_data
}) %>% 
  filter(NAME == "Lynchburg, VA Metro Area")

cpi <- read_excel("data/CPI_U_RS.xlsx") |> 
  rename(year = Year,
         priceindex = Index) |> 
  transform(year = as.numeric(year))

b19013_msa <- b19013_raw_msa %>% 
  mutate(across(.fns = ~str_remove_all(.x, "Alone Householder")),
         across(.fns = ~trimws(.x))) |> 
  select(year, NAME, race, estimate, moe) |> 
  transform(year = as.numeric(year),
            estimate = as.numeric(estimate))|> 
  left_join(cpi, by = 'year')|> 
  select(year, NAME, race, estimate, moe, priceindex) |> 
  transform(dollars21 = ((399/priceindex)*estimate)) |> 
  mutate(race = case_when(
    race == "Black Or African American" ~ "Black",
    race == "Two Or More Races Householder" ~ "Multiracial",
    race == "White Alone, Not Hispanic Or Latino Householder" ~ "White, non-Hispanic",
    race == "Hispanic Or Latino Householder" ~ "Hispanic or Latino",
    TRUE ~ race
  )) |> 
  filter(race %in% c("White, non-Hispanic", "Black", "Asian", "Multiracial", "Hispanic or Latino"))

write_rds(b19013_msa, "data/b19013_msa.rds")
```

```{r}
#| label: fig-inc-race-msa

b19013_msa <- read_rds("data/b19013_msa.rds") %>% 
  select(NAME, year, race, cdollars = estimate, dollars21) %>% 
  pivot_longer(4:5,
               names_to = "value",
               values_to = "estimate"
               ) %>% 
  mutate(value = case_when(
    value == "dollars21" ~ "2021 inflation-adjusted",
    value == "cdollars" ~ "Current dollars"))

race_inc_plot <- ggplot(b19013_msa,
       aes(x = year,
           y = estimate,
           color = race)) +
  geom_line(size = 0.5) +
  geom_point_interactive(
    aes(data_id = estimate,
        tooltip = dollar_format()(estimate)),
    size = 1) +
  theme_hfv() +
  scale_color_hfv() +
  facet_wrap(~value) +
  theme(legend.position = "bottom") +
  labs(title = "Median household income by race or ethnicity",
       subtitle = "From 2010 to 2021") +
  scale_y_continuous(labels = dollar_format())

if (knitr::is_html_output()) {
  
  girafe(ggobj = race_inc_plot,
         height_svg = 4)
} else {
  
  race_inc_plot
}

```

```{r}
#| label: inc-race-county
#| eval: false

years <- 2010:2021 

b19013 <- paste0("B19013", LETTERS[2:9])

b19013_defns <- load_variables(2020, "acs5") %>%
  filter(str_sub(name, end = 7) %in% b19013) %>%
  filter(str_detect(name, "PR") == FALSE)

concept_to_race <- function(x) {
  out <- x %>%
    str_remove_all("MEDIAN HOUSEHOLD INCOME IN THE PAST 12 MONTHS \\(\\IN 2020 INFLATION-ADJUSTED DOLLARS\\)\\ \\(|\\)") %>%
    str_to_title()
}

b19013_cleaned <- b19013_defns %>%
  mutate(race = concept_to_race(concept)) %>%
  separate(label, c("estimate", "medhhincome"), sep = "!!") %>%
  select(variable = name, medhhincome, race) %>%
  mutate(across(.fns = ~replace_na(.x, "All")),
          across(.fns = ~str_remove_all(.x, ":")))

b19013_raw <- map_dfr(b19013, function(tb) {
  yearly_data <- map_dfr(years, function(yr) {
    acs_pull <- get_acs(
      geography = "county",
      table = tb,
      year = yr,
      state = "VA"
    ) %>%
    left_join(b19013_cleaned, by = "variable")
    acs_rearranged <- acs_pull %>%
      mutate(year = yr) %>%
      select(variable, year, NAME, GEOID, race, medhhincome,
             estimate, moe)
    acs_rearranged
  })
  yearly_data
})

cpi <- read_excel("data/CPI_U_RS.xlsx") |> 
  rename(year = Year,
         priceindex = Index) |> 
  transform(year = as.numeric(year))

b19013_data <- b19013_raw |> 
  subset(GEOID %in% cv) |> 
  mutate(across(.fns = ~str_remove_all(.x, "Alone Householder")),
         across(.fns = ~str_remove_all(.x, ", Virginia")),
         across(.fns = ~trimws(.x))) |> 
  select(year, NAME, GEOID, race, estimate, moe) |> 
  transform(year = as.numeric(year),
            estimate = as.numeric(estimate))|> 
  left_join(cpi, by = 'year')|> 
  select(year, NAME, GEOID, race, estimate, moe, priceindex) |> 
  transform(dollars21 = ((399/priceindex)*estimate)) |> 
  mutate(race = case_when(
    race == "Black Or African American" ~ "Black",
    race == "Two Or More Races Householder" ~ "Multiracial",
    race == "White Alone, Not Hispanic Or Latino Householder" ~ "White, non-Hispanic",
    race == "Hispanic Or Latino Householder" ~ "Hispanic or Latino",
    TRUE ~ race
  )) |> 
  filter(race %in% c("White, non-Hispanic", "Black", "Asian", "Multiracial", "Hispanic or Latino"))

write_rds(b19013_data, "data/b19013_data.rds")
```

## Housing stock

```{r}
#| label: structure-data
#| eval: false

# Table B25127: Tenure by year structure built by units in structure

years <- 2010:2021

b25127_vars <- load_variables(2021, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25127")

b25127_raw <- map_dfr(years, function(yr){
  b25127_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B25127",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |> 
    mutate(year = yr)
})

b25127_raw <- b25127_raw |> 
  subset(GEOID %in% cv) 

b25127_vars_cleaned <- b25127_vars |> 
  separate(label, into = c("est", "total", "tenure", "yrbuilt", "structure"), sep = "!!") |> 
  select(variable = name, tenure, yrbuilt, structure) |> 
  drop_na() |> 
  mutate(across(.fns = ~str_remove_all(.x, ":"))) |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  ))

b25127_raw <- b25127_raw |> 
  right_join(b25127_vars_cleaned, by = "variable") |> 
  select(NAME, GEOID, year, tenure, yrbuilt, structure, estimate, moe)

b25127_data <- b25127_raw |> 
  mutate(NAME = str_remove_all(NAME, ", Virginia"),
         yrbuilt = str_remove_all(yrbuilt, "Built "),
         NAME = case_when(
           NAME == "Bedford city" ~ "Bedford County", # Aggregate Bedford City and Bedford County
           TRUE ~ NAME)) |>
  select(NAME, GEOID, year, tenure, yrbuilt, structure, estimate) |> 
  mutate(structure = case_when(
    structure == "1, detached  or attached" ~ "Single-family",
    structure == "2 to 4" ~ "2 to 4 units",
    structure == "5 to 19" ~ "5 to 19 units",
    structure == "20 to 49" ~ "20 or more units",
    structure == "50 or more" ~ "20 or more units",
    TRUE ~ "Other"
  )) 

write_rds(b25127_data, "data/b25127_data.rds")

```

```{r}
#| label: fig-structure

b25127_data <- read_rds("data/b25127_data.rds")

cv_structure <- b25127_data %>% 
  group_by(year, tenure, structure) %>% 
  summarise(estimate = sum(estimate)) %>% 
  group_by(year) %>% 
  mutate(pct = estimate/sum(estimate)) %>% 
  mutate(year = as.character(year))

cv_structure_plot <- ggplot(cv_structure,
       aes(x = year,
           y = pct,
           fill = structure,
           data_id = pct,
           tooltip = percent_format()(pct))) +
  geom_col(position = "stack") +
  geom_col_interactive() +
  facet_wrap(~tenure) +
  scale_y_continuous(labels = percent_format()) +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Housing stock by structure type",
       subtitle = "Percent of total occupied housing from 2010 to 2021",
       caption = "**Source:** U.S. Census Bureau, 5-Year ACS, Table B25127.") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_structure_plot,
         height_svg = 4)
  
} else {
  
  cv_structure_plot
}

```

```{r}
#| label: fig-yrbuilt

cv_yrbuilt <- b25127_data %>% 
  filter(year == 2021) %>% 
  group_by(tenure, yrbuilt) %>% 
  summarise(estimate = sum(estimate))

cv_yrbuilt_plot <- ggplot(cv_yrbuilt,
       aes(x = yrbuilt,
           y = estimate,
           fill = yrbuilt,
           data_id = estimate,
           tooltip = label_comma()(estimate))) +
  geom_col() +
  geom_col_interactive() +
  facet_wrap(~tenure) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Housing stock by year built and tenure",
       caption = "U.S. Census Bureau, 5-year ACS, Table B25127.") + 
  scale_x_discrete(labels = function(x)str_wrap(x, width = 9)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_yrbuilt_plot,
         height_svg = 4)
} else {
  
  cv_yrbuilt_plot
}

```

```{r}
#| label: beds-data
#| eval: false

years <- 2010:2021

b25042_vars <- load_variables(2021, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25042")

b25042_raw <- map_dfr(years, function(yr) {
  b25042_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B25042",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |> 
    mutate(year = yr)
})

b25042_raw <- b25042_raw %>% 
  subset(GEOID %in% cv)

b25042_vars_cleaned <- b25042_vars |> 
  separate(label, into = c("est", "total", "tenure", "br"), sep = "!!") |> 
  select(variable = name, tenure, br) |> 
  drop_na() |> 
  mutate(across(.fns = ~str_remove_all(.x, ":"))) |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  ))

b25042_raw <- b25042_raw |> 
  right_join(b25042_vars_cleaned, by = "variable") |> 
  select(NAME, GEOID, year, tenure, br, estimate)

b25042_data <- b25042_raw |> 
  mutate(NAME = str_remove_all(NAME, ", Virginia"),
         NAME = case_when(
           NAME == "Bedford city" ~ "Bedford County", # Aggregate Bedford City and Bedford County
           TRUE ~ NAME)) |>
  select(NAME, GEOID, year, tenure, br, estimate) 


write_rds(b25042_data, "data/b25042_data.rds")
```

```{r}
#| label: fig-beds

b25042_data <- read_rds("data/b25042_data.rds")

cv_beds <- b25042_data %>% 
  mutate(br = case_when(
    br == "4 bedrooms" ~ "4 or more bedrooms",
    br == "5 or more bedrooms" ~ "4 or more bedrooms",
    TRUE ~ br
  )) %>% 
  group_by(year, tenure, br) %>% 
  summarise(estimate = sum(estimate)) %>% 
  pivot_wider(
    names_from = year,
    values_from = estimate
  ) %>% 
  mutate(chg = `2021` - `2010`)


cv_beds_plot <- ggplot(cv_beds,
       aes(x = br,
           y = chg,
           fill = br,
           data_id = chg,
           tooltip = label_comma()(chg))) +
  geom_col(position = "stack") +
  geom_col_interactive() +
  facet_wrap(~tenure) +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Change in bedrooms by tenure",
       subtitle = "From 2010 to 2021") +
  scale_x_discrete(labels = function(x)str_wrap(x, width = 8)) +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus"))
  
if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_beds_plot,
         height_svg = 4)
} else {
  
  cv_beds_plot
}

```

```{r}
#| label: fig-mhc-map

mhc <- st_read("https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Mobile_Home_Parks/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson", quiet = TRUE) |> 
  filter(STATE == "VA") %>% 
  filter(str_detect(COUNTY, "AMHERST|APPOMATTOX|BEDFORD|CAMPBELL|LYNCHBURG"))

library(leaflet)

leaflet(mhc) |> 
  addTiles() |> 
  addCircleMarkers()


```


```{r}
#| label: bps-data
#| eval: false

bps_years <- 2000:2021

header_rows <- read_csv("https://www2.census.gov/econ/bps/County/co2020a.txt", 
                        col_names = FALSE,
                        n_max = 2)

column_names <- header_rows %>%
  select(X1:X18) %>%
  t() %>%
  as_tibble() %>%
  mutate(group = rep(1:6, each = 3)) %>%
  group_by(group) %>%
  fill(V1, .direction = "updown") %>%
  mutate(names = paste0(V1, ": ", V2)) %>%
  pull(names)

library(glue)

cbps_raw <- map_df(bps_years, ~{
  raw <- read_csv(glue("https://www2.census.gov/econ/bps/County/co{.x}a.txt"), skip = 2,
                    col_names = FALSE) %>%
    select(X1:X18) %>%
    set_names(column_names)
  
  raw
  
})


# Read in latest 2022 cumulative data (through November)

cbps_ytd <- read_csv("https://www2.census.gov/econ/bps/County/co2211y.txt", 
                     col_names = FALSE,
                     skip = 2) |> 
  select(X1:X18) |> 
  set_names(column_names) |> 
  mutate(`Survey: Date` = 2022)

cbps_data <- cbps_raw %>% 
  bind_rows(cbps_ytd) |> 
  mutate(year = `Survey: Date`,
         GEOID = paste0(`FIPS: State`, `FIPS: County`)) %>%
  select(`1-unit: Bldgs`:GEOID) %>%
  filter(GEOID %in% cv) %>%
  pivot_longer(`1-unit: Bldgs`:`5+ units: Value`,
               names_to = "type",
               values_to = "value") %>%
  separate(type, into = c("Type", "col"), sep = ": ") %>%
  pivot_wider(names_from = col,
              values_from = value) %>%
  rename_with(tolower, Type:Value) %>% 
  select(GEOID, year, type:units)

cv_cbps <- cbps_data %>% 
  mutate(GEOID = case_when(
    GEOID == "51515" ~ "51019",
    TRUE ~ GEOID)) %>% 
  left_join(counties("VA", year = 2021), by = "GEOID") %>% 
  select(GEOID, NAMELSAD, year, type, units) %>% 
  group_by(GEOID, NAMELSAD, year, type) %>% 
  summarise(units = sum(units))


write_rds(cv_cbps, "data/cv_cbps.rds")
```

```{r}
#| label: fig-permits

cv_cbps <- read_rds("data/cv_cbps.rds") %>% 
  group_by(year, type) %>% 
  summarise(units = sum(units))

cv_cbps_plot <- ggplot(cv_cbps,
       aes(x = year,
           y = units,
           fill = forcats::fct_rev(type),
           data_id = units,
           tooltip = label_comma()(units))) +
  geom_col(position = "stack") +
  geom_col_interactive() +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "New residential building permits",
       subtitle = "From 2000 through November 2022")

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_cbps_plot,
         height_svg = 4)
} else {
  
  cv_cbps_plot
}


```

## Homeownership market

```{r}
#| label: ho-rate-data

# Read in tenure data via Table B25003.

b25003_data <- read_rds("data/b25003_data.rds")

# Aggregate data to the regional level.

ho_rate <- b25003_data %>%
  select(year, tenure, estimate) %>% 
  group_by(year, tenure) %>% 
  summarise(estimate = sum(estimate))

# Calculate the percent of total by year and tenure.

cv_ho <- ho_rate %>% 
  group_by(year) %>% 
  mutate(pct = estimate/sum(estimate)) %>%  
  filter(tenure == "Homeowner") # Retain only homeowners to get the homeownership rate.

cv_ho_plot <- ggplot(cv_ho,
       aes(x = year, 
           y = pct)) +
  geom_line(size = 0.5) +
  geom_point_interactive(
    aes(data_id = estimate,
        tooltip = dollar_format()(estimate)),
    size = 1) +
  theme_hfv() +
  scale_color_hfv() +
  scale_y_continuous(labels = percent_format(), limits = c(.65,.75)) +
  scale_x_continuous(breaks = seq(from = 2010, to = 2021, by = 1)) +
  labs(title = "Homeownership rate",
       subtitle = "From 2010 to 2021",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25003.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_ho_plot,
         height_svg = 4)
} else {
  
  cv_ho_plot
}

```

```{r}
#| label: fig-mls-price

# Utilize Lynchburg MLS data. Need median sales price for the entire region from 2000 to 2022 or at least 2010. Monthly median sales price is preferred.



```


```{r}
#| label: fig-mls-sales

# Utilzie Lynchburg MLS data. Need mumber of closed sales for the entire region from 2000 to 2022 or at least 2010. Monthly number is preferred.


```

```{r}
#| label: fig-mls-ratio

# Utilize Lynchburg MLS data. Need close-to-list ratio for the entire region from 2000 to 2022 or at least 2010. Monthly ratio is preferred.

```

## Rental market

```{r}
#| label: rent

# Utilize CoStar properties search for Multi-Family. Filter by MSA - Lynchburg - VA Market. Then review Analytics and pull QUARTERLY Market Asking Rent Per Unit for entire region. Do not bring in Forecast and set History to All.

# Download data for current rent and inflation-adjusted rent into a single csv - name cvpdc_rent.csv

# Note the download date of data here: 1/1/2023

cvpdc_rent <- read_csv("data/cvpdc_rent.csv") %>%
  filter(!str_detect(period, "QTD")) %>% 
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) %>% 
  pivot_longer(2:3,
               names_to = "value",
               values_to = "estimate") %>% 
  arrange(desc(value))

ggplot(cvpdc_rent,
       aes(x = period,
           y = estimate,
           linetype = value,
           color = value)) + 
  geom_line(size = 0.5) +
  geom_point_interactive(
    aes(data_id = estimate,
        tooltip = dollar_format()(estimate)),
    size = 1) +
  scale_y_continuous(labels = dollar_format(), limits = c(500,1500)) +
  scale_linetype_manual(values = c(2,1)) +
  theme_hfv() +
  scale_color_hfv() +
  theme(legend.position = "bottom") + 
  labs(title = "Average quarterly market asking rent")



```

```{r}
#| label: vacancy

# Utilize CoStar properties search for Multi-Family. Filter by Market for Lynchburg MSA- VA Market. Then review Analytics and pull QUARTERLY Vacancy Rate for entire region. Do not bring in Forecast and set History to All.

# Download data into csv - name cvpdc_vacancy.csv

# Note the download date of data here: 

cvpdc_vacancy <- read_csv("data/cvpdc_vacancy.csv") %>% 
  select(period, vacancy = overall_vacancy) %>% 
  filter(!str_detect(period, "QTD")) %>% 
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1))

ggplot(cvpdc_vacancy,
       aes(x = period,
           y = vacancy)) +
  geom_line(size = 1) +
  theme_hfv() +
  scale_color_hfv() +
  labs(title = "Rental vacancy rate")


```

## Affordability

```{r}
#| label: cb-data

years <- 2012:2019

sumlev <- "050"

dir.create(glue("data/{sumlev}"))

walk(years, ~{
  url <- glue("https://www.huduser.gov/PORTAL/datasets/cp/{.x - 4}thru{.x}-{sumlev}-csv.zip")
  file <- basename(url)
  path <- file.path("data", sumlev, file)
  if (!file.exists(path)) {
    GET(url, write_disk(path, overwrite = TRUE), progress(type = "down"))
  }
  print(glue("Unzipping {.x}..."))
  unzip(path, exdir = file.path("data", sumlev, .x))
})

# Tables to get
tables <- c(7, 9, paste0(18, LETTERS[1:3]))

# Go through and write out the various tables
walk(tables, function(table) {
  mytable <- purrr::map_df(years, function(year) {
    # Identify the year-specific folder
    path <- file.path("data", "050", year)
    # Find the file - it may be buried so use recursive = TRUE
    file <- list.files(path, pattern = glue("Table{table}.csv"), recursive = TRUE)
    # Read in the file quietly
    raw <- read_csv(file.path(path, file), col_types = cols())
    # Clean it up
    cleaned <- raw |>
      clean_names() |>
      mutate(fips = substr(geoid, 8, 12)) |>
      separate(name, into = c("county", "state"), sep = ",") |>
      filter(st == "51") |>
      pivot_longer(starts_with("T"),
                   names_to = "code",
                   values_to = "value") |>
      mutate(id = str_extract(code, "\\d+$"),
             type = str_extract(code, "est|moe")) |>
      select(-code) |>
      pivot_wider(names_from = type, values_from = value) |>
      rename(Estimate = est, MOE = moe) |>
      mutate(Code := glue("T{table}_est{id}"),
             Year = year) |>
      select(Code, Year, Estimate, MOE, everything(), -id) |>
      mutate(fips = case_when(
        fips == "51515" ~ "51019",
        TRUE ~ fips
      )) |>
      mutate(county = case_when(
        county == "Bedford city" ~ "Bedford County",
        TRUE ~ county
      )) |>
      subset(fips %in% cv)
    
  
    # Account for different data dictionaries
    # Find the data dictionary in the appropriate folder
    dict_path <- list.files(path, pattern = "dictionary", recursive = TRUE, full.names = TRUE)
    
    # Read in the data dictionary and merge
    dict <- read_excel(dict_path,
                       sheet = glue("Table {table}"))
    cleaned_with_dict <- cleaned %>%
      left_join(dict, by = c("Code" = "Column Name"))
    cleaned_with_dict
  })
  file_name <- glue("Table{table}_2012to2019.csv")
  message(glue("Writing file {file_name}..."))
  write_csv(mytable, glue("data/{file_name}"))
})

```



```{r}
#| label: cb7-clean
#| eval: false

cb_7 <- read_csv("data/Table7_2012to2019.csv") %>% 
  clean_names() %>% 
  filter(line_type == "Detail") %>% 
  select(year, estimate, moe, county, fips, tenure, household_income, household_type, cost_burden) %>% 
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  )) %>% 
  mutate(household_income = case_when(
    household_income == "household income is less than or equal to 30% of HAMFI" ~ "30% AMI or less",
    household_income == "household income is greater than 30% but less than or equal to 50% of HAMFI" ~ "31 to 50% AMI",
    household_income == "household income is greater than 50% but less than or equal to 80% of HAMFI" ~ "51 to 80% AMI",
    household_income == "household income is greater than 80% but less than or equal to 100% of HAMFI" ~ "81 to 100% AMI",
    household_income == "household income is greater than 100% of HAMFI" ~ "101% AMI or greater"
  )) %>% 
  mutate(household_type = case_when(
    household_type == "household type is elderly family (2 persons, with either or both age 62 or over)" ~ "Elderly family",
    household_type == "household type is small family (2 persons, neither person 62 years or over, or 3 or 4 persons)" ~ "Small family",
    household_type == "household type is large family (5 or more persons)" ~ "Large family",
    household_type == "household type is elderly non-family" ~ "Elderly non-family",
    household_type == "other household type (non-elderly non-family)" ~ "Non-elderly non-family"
  ))  %>% 
  mutate(cost_burden = case_when(
    cost_burden == "housing cost burden is less than or equal to 30%" ~ "Not cost-burdened",
    cost_burden == "housing cost burden is greater than 30% but less than or equal to 50%" ~ "Cost-burdened",
    cost_burden == "housing cost burden is greater than 50%" ~ "Severely cost-burdened",
    cost_burden == "housing cost burden not computed (household has no/negative income)" ~ "No or negative income"
  )) %>% 
  mutate(cb_group = case_when(
    cost_burden == "Cost-burdened" ~ "Cost-burdened",
    cost_burden == "Severely cost-burdened" ~ "Cost-burdened",
    TRUE ~ cost_burden
  ))

write_rds(cb_7, "data/cb_7.rds")

```

```{r}
#| label: fig-cb-total


cb_7 <- read_rds("data/cb_7.rds")
  
cbhh_region <- cb_7 %>% 
  group_by(year, household_income, cb_group) %>% 
  summarise(estimate = sum(estimate)) %>% 
  group_by(year, household_income) %>% 
  mutate(pct = estimate/sum(estimate)) %>% 
  filter(cb_group == "Cost-burdened")

cbhh_region$household_income <- factor(cbhh_region$household_income,
                                       levels = c("30% AMI or less", "31 to 50% AMI", "51 to 80% AMI", "81 to 100% AMI", "101% AMI or greater"))

cbhh_region_plot <- ggplot(cbhh_region,
       aes(x = year, 
           y = pct, 
           fill = household_income,
           data_id = pct,
           tooltip = percent_format()(pct))) +
  geom_col(position = "stack") +
  geom_col_interactive(position = "stack", size = 2) +
  theme_hfv() +
  scale_fill_hfv() +
  facet_wrap(~household_income, nrow = 1) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Cost-burdened households by household income",
       subtitle = "Percent of cost-burdened households by household income from 2012 to 2019",
       caption = "**Source:** U.S Department of Housing and Urban Development,<br>Comprehensive Housing Affordability Strategy (CHAS), Table 7.") +
  add_zero_line(axis = "y") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))


if (knitr::is_html_output()) {
  
  girafe(ggobj = cbhh_region_plot,
         height_svg = 4) 
  
} else {
  
  cbhh_region_plot
  
  }


```

```{r}
#| label: fig-cb-change
#| fig-cap: "Change in cost-burdened households by tenure"


cb_7 <- read_rds("data/cb_7.rds")
  
cb_change <- cb_7 %>% 
  filter(fips %in% cv,
         cb_group == "Cost-burdened") |> 
  group_by(year, tenure) |> 
  summarise(estimate = sum(estimate)) |> 
  group_by(tenure) |> 
  mutate(change = estimate - lag(estimate)) |> 
  drop_na() |> 
  mutate(run_diff = cumsum(change))
  
cb_diff <- ggplot(cb_change,
       aes(x = year, 
           y = run_diff, 
           fill = tenure,
           data_id = run_diff,
           tooltip = run_diff)) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge", size = 2) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Cumulative change in cost-burdened households by tenure",
       subtitle = "2012 to 2019",
       fill = "Tenure",
       caption = "**Source:** U.S Department of Housing and Urban Development,<br>Comprehensive Housing Affordability Strategy (CHAS), Table 7.") +
  theme(legend.position = "right") +
  add_zero_line(axis = "y")


if (knitr::is_html_output()) {
  
  girafe(ggobj = cb_diff,
         height_svg = 4) 
  
} else {
  
  cb_diff
  
  }


```


```{r}
#| label: cb9-clean
#| eval: false

cb_9 <- read_csv("data/Table9_2012to2019.csv") %>% 
  clean_names() %>% 
  filter(line_type == "Detail")  %>% 
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  )) %>% 
  mutate(race = case_when(
    race_ethnicity == "White alone, non-Hispanic" ~ "White, non-Hispanic",
    race_ethnicity == "Black or African-American alone, non-Hispanic" ~ "Black",
    race_ethnicity == "Asian alone, non-Hispanic" ~ "Asian",
    race_ethnicity == "Hispanic, any race" ~ "Hispanic or Latino",
    TRUE ~ "Another race, including multiracial"
  ))   %>% 
  mutate(cost_burden = case_when(
    cost_burden == "less than or equal to 30%" ~ "Not cost-burdened",
    cost_burden == "greater than 30% but less than or equal to 50%" ~ "Cost-burdened",
    cost_burden == "greater than 50%" ~ "Severely cost-burdened",
    cost_burden == "not computed (no/negative income)" ~ "No or negative income"
  )) %>% 
  mutate(cb_group = case_when(
    cost_burden == "Cost-burdened" ~ "Cost-burdened",
    cost_burden == "Severely cost-burdened" ~ "Cost-burdened",
    TRUE ~ cost_burden
  )) %>% 
  select(county, year, race, cost_burden, cb_group, estimate)

write_rds(cb_9, "data/cb_9.rds")

```

```{r}
#| label: fig-cb-race
#| fig-cap: "Percent of cost-burdened households by race and ethnicity"

cb_9 <- read_rds("data/cb_9.rds")

cv_cb9 <- cb_9 %>% 
  group_by(year, race, cb_group) %>% 
  summarise(estimate = sum(estimate)) %>% 
  group_by(year, race) %>% 
  mutate(pct = estimate/sum(estimate)) %>% 
  filter(cb_group == "Cost-burdened") %>% 
  group_by(race) %>% 
  mutate(change = pct - lag(pct)) %>% 
  mutate(year = as.character(year))

cv_cb9_plot <- ggplot(cv_cb9,
       aes(x = year, 
           y = pct,
           fill = race,
           data_id = pct,
           tooltip = percent_format()(pct))) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  facet_wrap(~race, nrow = 1, labeller = label_wrap_gen()) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_hfv() +
  theme_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        strip.text = element_text(size = 8)) +
  labs(title = "Cost burden by race and ethnicity",
       subtitle = "Percent of cost-burdened households",
       caption = "**Source:** U.S. Department of Housing and Urban Development, Comprehensive Housing Affordability Strategy (CHAS), Table 9.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_cb9_plot,
         height_svg = 4)
  
} else {
  
  cv_cb9_plot
  
}

```
