# County Housing Market Assessment

The following provides a locality-level analysis of major trends impacting housing within Central Virginia Planning District region. All data has been disaggregated to show the differences between localities

```{r}
#| label: setup

library(tidyverse)
library(tigris)
library(hdatools)
library(ggiraph)
library(scales)
library(ggtext)

cv <- c("51009", "51011", "51019", "51031", "51680", "51515") # CVPDC FIPS codes

cv_names <- list_counties("VA") %>% 
  mutate(GEOID = paste0("51", county_code))

```

## Takeaways

## Population trends

```{r}
#| label: pop

pop_data <- read_rds("data/pop_data.rds")

local_pop <- pop_data %>% 
  left_join(cv_names, by = "GEOID") %>% 
  select(year, county, counttype, value) 

local_pop$county <- factor(local_pop$county, levels = c("Lynchburg city", "Bedford", "Campbell", "Amherst", "Appomattox"))

ggplot(local_pop,
       aes(x = year,
           y = value,
           fill = counttype)) +
  geom_col() +
  facet_wrap(~county, nrow = 1) +
  theme_hfv() +
  scale_color_hfv() +
  theme(axis.text.x = element_text(angle = 90))


```

```{r}
#| label: pop-change

comp_data <- read_rds("data/comp_data.rds")

# Create group bar chart showing population change by locality.

ggplot(comp_data,
       aes(x = year, 
           y = value,
           fill = component)) +
  geom_col(position = "stack") +
  facet_wrap(~NAMELSAD, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(size = 6, angle = 90),
        legend.position = "bottom") +
  labs(title = "Components of population change",
       caption = "**Source:** U.S. Census Bureau, Population Estimates Program.")



```

```{r}
#| label: fig-projections


projections <- read_csv("data/uva_proj.csv") %>% 
  subset(FIPS %in% cv)


```

## Household trends

```{r}
#| label: tenure

b25003_data <- read_rds("data/b25003_data.rds")


```

```{r}
#| label: age

b25007_data <- read_rds("data/b25007_data.rds")


```

```{r}
#| label: type

b11001_data <- read_rds("data/b11001_data.rds")


```

```{r}
#| label: size

b25009_data <- read_rds("data/b25009_data.rds")


```

```{r}
#| label: size

b09020_data <- read_rds("data/b09020_data.rds")


```

```{r}
#| label: subfam

b11013_data <- read_rds("data/b11013_data.rds")


```

```{r}
#| label: achild

b09021_data <- read_rds("data/b09021_data.rds")



```


## Economic trends

```{r}
#| label: inc-dist

b25118_data <- read_rds("data/b25118_data.rds")


```

```{r}
#| label: med-inc

b25118_data <- read_rds("data/b25119_cpi.rds")


```

```{r}
#| label: race-inc

b19013_data <- read_rds("data/b19013_data.rds")


```


## Housing stock

```{r}
#| label: structure

b25127_data <- read_rds("data/b25127_data.rds")


```

```{r}
#| label: mhc


mhc <- st_read("https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Mobile_Home_Parks/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson", quiet = TRUE) |> 
  filter(STATE == "VA") %>% 
  filter(str_detect(COUNTY, "AMHERST|APPOMATTOX|BEDFORD|CAMPBELL|LYNCHBURG"))



```

```{r}
#| label: bps

cv_cbps <- read_rds("data/cv_cbps.rds")

```

## Homeownership market

```{r}
#| label: ho

b25003_data <- read_rds("data/b25003_data.rds")

```


## Rental market

```{r}
#| label: rent

# Utilize CoStar properties search for Multi-Family. Filter by County for Lynchburg - VA Market. Then review Analytics and pull QUARTERLY Market Asking Rent Per Unit for entire region. Do not bring in Forecast and set History to All.

# Download data for current rent and inflation-adjusted rent into a single csv - name [locality_abbv]_rent.csv --- for example amherst_rent.csv or lburg_rent.csv

# Merge all data into a single csv named - local_rent.csv

# Note the download date of data here: 






```

```{r}
#| label: vacancy

# Utilize CoStar properties search for Multi-Family. Filter by County for Lynchburg - VA Market. Then review Analytics and pull QUARTERLY Vacancy Rate for entire region. Do not bring in Forecast and set History to All.

# Download data into csv - name [locality_abbv]_vacancy.csv --- for example amherst_vacancy.csv or lburg_vacancy.csv

# Note the download date of data here: 




```




