# County Housing Market Assessment

The following provides a locality-level analysis of major trends impacting housing within Central Virginia Planning District region. All data has been disaggregated to show the differences between localities

```{r}
#| label: setup

library(tidyverse)
library(tigris)
library(hdatools)
library(ggiraph)
library(scales)
library(ggtext)

cv <- c("51009", "51011", "51019", "51031", "51680", "51515") # CVPDC FIPS codes

cv_names <- list_counties("VA") %>% 
  mutate(GEOID = paste0("51", county_code))

```

## Takeaways

## Population trends

```{r}
#| label: pop-2020

pop_data <- read_rds("data/pop_data.rds")

local_pop <- pop_data %>% 
  left_join(cv_names, by = "GEOID") %>% 
  select(year, county, counttype, value) %>% 
  group_by(county) %>% 
  mutate(pchg = (value - first(value))/first(value)) %>% 
  filter(year == 2020) %>% 
  select(county, value, pchg)


local_pop_plot <- ggplot(local_pop,
       aes(x = reorder(county, -value),
           y = value,
           fill = county)) +
  geom_col() +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "2020 Census population",
       caption = "**Source:** U.S. Census Bureau, Decennial Census.") +
  scale_y_continuous(labels = label_comma(), limits = c(0, 100000))

if (knitr::is_html_output()) {
  
  girage(ggobj = local_pop_plot,
         height_svg = 4)
  
} else {
  
  local_pop_plot
  
}


```

```{r}

local_growth_plot <- ggplot(local_pop,
       aes(x = reorder(county, -pchg),
           y = pchg,
           fill = county)) +
  geom_col() +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = percent_format(), limits = c(-.05, .2)) +
  labs(title = "Population growth",
       subtitle = "Percent change in population from 2010",
       caption = "**Source:** U.S. Census Bureau, Population Estimates Program.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_growth_plot,
         height_svg = 4)
  
} else {
  
  local_growth_plot
  
}


```

```{r}
#| label: fig-local-chg

local_change <- read_rds("data/comp_data.rds")

# Create group bar chart showing population change by locality.

local_change_plot <- ggplot(local_change,
       aes(x = year, 
           y = value,
           fill = component)) +
  geom_col(position = "stack") +
  facet_wrap(~NAMELSAD, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(size = 6, angle = 90),
        legend.position = "bottom") +
  labs(title = "Components of population change",
       caption = "**Source:** U.S. Census Bureau, Population Estimates Program.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_change_plot,
         height_svg = 4)
  
} else {
  
  local_change_plot
  
}


```

```{r}
#| label: fig-projections


projections <- read_csv("data/uva_proj.csv") %>% 
  subset(FIPS %in% cv) %>% 
  mutate(Difference = as.numeric(Difference))

local_proj_plot <- ggplot(projections,
       aes(x = reorder(NAME, -Difference),
           y = Difference,
           fill = NAME)) +
  geom_col() +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Projected growth",
       subtitle = "Change from 2030 to 2050",
       caption = "University of Virginia, Weldon Cooper Center for Public Service.") +
  scale_y_continuous(labels = label_comma())

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_proj_plot,
         height_svg = 4)
  
} else {
  
  local_proj_plot 
  
}
  
```

## Household trends

```{r}
#| label: tenure

b25003_data <- read_rds("data/b25003_data.rds")

local_tenure <- b25003_data %>%
  select(county = NAME, year, tenure, estimate) %>% 
  group_by(county, year, tenure) %>% 
  summarise(estimate = sum(estimate)) %>% 
  group_by(county, tenure) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) %>% 
  mutate(year = as.character(year))


local_tenure_plot <- ggplot(local_tenure,
       aes(x = year,
           y = run_diff,
           fill = tenure,
           data_id = run_diff,
           tooltip )) +
  geom_col(position = "dodge") +
  facet_wrap(~county, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, size = 8)) +
  labs(title = "Cumulative change in households by tenure",
       subtitle = "From 2010 to 2021",
       caption = "U.S. Census Bureau, ACS 5-Year Estimates, Table B25003.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_tenure_plot,
         height_svg = 4)
  
} else {
  
  local_tenure_plot
}
  

```

```{r}
#| label: age

b25007_data <- read_rds("data/b25007_data.rds")
  
local_age <- b25007_data %>% 
  group_by(county = NAME, year, tenure, age) %>% 
  summarise(estimate = sum(estimate)) %>% 
  filter(year == 2010 | year == 2021) %>% 
  group_by(county, year, tenure) %>% 
  mutate(pct = estimate/sum(estimate)) %>% 
  mutate(year = as.character(year))

local_age_plot <- ggplot(local_age,
       aes(x = year, 
           y = pct,
           fill = age,
           data_id = pct,
           tooltip = percent_format()(pct)))  +
  geom_col(position = "stack") +
  facet_grid(tenure~county) +
  geom_col_interactive(size = 2) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = percent_format()) +
  theme(legend.position = "bottom") +
  labs(title = "Percent of households by age and tenure",
       subtitle = "In 2010 and 2021",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25007.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_age_plot,
         height_svg = 4)
  
} else {
  
  local_age_plot
}

```

```{r}
#| label: type

b11001_data <- read_rds("data/b11001_data.rds")


```

```{r}
#| label: size

b25009_data <- read_rds("data/b25009_data.rds")

local_size <- b25009_data %>% 
  select(county = NAME, year, tenure, size, estimate) %>% 
  group_by(county, year, tenure, size) %>% 
  summarise(estimate = sum(estimate))  %>%
  group_by(county, tenure, size) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) %>% 
  mutate(year = as.character(year)) %>% 
  filter(year == "2021")


local_size_plot <- ggplot(local_size,
       aes(x = county,
           y = run_diff,
           fill = tenure,
           data_id = run_diff,
           tooltip )) +
  geom_col(position = "dodge") +
  facet_wrap(~size, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, size = 8)) +
  labs(title = "Cumulative change in households by size and tenure",
       subtitle = "From 2010 to 2021",
       caption = "U.S. Census Bureau, ACS 5-Year Estimates, Table B25009.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_size_plot,
         height_svg = 4)
  
} else {
  
  local_size_plot
}
  


```

```{r}
#| label: senior

b09020_data <- read_rds("data/b09020_data.rds")

local_senior <- b09020_data %>% 
  select(county = NAME, year, family, relationship, estimate) %>% 
  group_by(county, year, family, relationship) %>% 
  summarise(estimate = sum(estimate)) %>% 
  

```

```{r}
#| label: subfam

b11013_data <- read_rds("data/b11013_data.rds")


```

```{r}
#| label: achild

b09021_data <- read_rds("data/b09021_data.rds")



```


## Economic trends

```{r}
#| label: inc-dist

b25118_data <- read_rds("data/b25118_data.rds")


```

```{r}
#| label: med-inc

b25118_data <- read_rds("data/b25119_cpi.rds")


```

```{r}
#| label: race-inc

b19013_data <- read_rds("data/b19013_data.rds")


```


## Housing stock

```{r}
#| label: structure

b25127_data <- read_rds("data/b25127_data.rds")


```

```{r}
#| label: mhc


mhc <- st_read("https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Mobile_Home_Parks/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson", quiet = TRUE) |> 
  filter(STATE == "VA") %>% 
  filter(str_detect(COUNTY, "AMHERST|APPOMATTOX|BEDFORD|CAMPBELL|LYNCHBURG"))



```

```{r}
#| label: fig-mhc-county


cv_mhc <- mhc  %>% 
  group_by(COUNTY, SIZE) %>% 
  summarise(count = n_distinct(OBJECTID)) %>% 
  st_drop_geometry() %>% 
  mutate(COUNTY = str_to_sentence(COUNTY))

ggplot(cv_mhc,
       aes(x = reorder(COUNTY,count),
           y = count,
           fill = SIZE)) +
  geom_col(position = "stack") +
  theme_hfv() +
  scale_fill_hfv()


```

```{r}
#| label: bps

cv_cbps <- read_rds("data/cv_cbps.rds")

```

## Homeownership market

```{r}
#| label: ho

b25003_data <- read_rds("data/b25003_data.rds")

```


## Rental market

```{r}
#| label: rent

# Utilize CoStar properties search for Multi-Family. Filter by County for Lynchburg - VA Market. Then review Analytics and pull QUARTERLY Market Asking Rent Per Unit for entire region. Do not bring in Forecast and set History to All.

# Download data for current rent and inflation-adjusted rent into a single csv - name [locality_abbv]_rent.csv --- for example amherst_rent.csv or lburg_rent.csv

# Merge all data into a single csv named - local_rent.csv

# Note the download date of data here: 






```

```{r}
#| label: vacancy

# Utilize CoStar properties search for Multi-Family. Filter by County for Lynchburg - VA Market. Then review Analytics and pull QUARTERLY Vacancy Rate for entire region. Do not bring in Forecast and set History to All.

# Download data into csv - name [locality_abbv]_vacancy.csv --- for example amherst_vacancy.csv or lburg_vacancy.csv

# Note the download date of data here: 




```




