# County Housing Market Assessment

The following provides a county-level analysis of major trends impacting housing within Central Virginia Planning District region. All data has been disaggregated to show the differences between localities.

```{r}
#| label: setup

library(tidyverse)
library(tigris)
library(hdatools)
library(ggiraph)
library(scales)
library(ggtext)
library(tidytext)
library(janitor)

cnty <- c("51009", "51011", "51019", "51031", "51515") # County FIPS codes

cv_names <- list_counties("VA") %>% 
  mutate(GEOID = paste0("51", county_code))

```

## Takeaways

## Population trends

```{r}
#| label: pop-2020

# Read in population data and filter for only the counties using the county list.

pop_data <- read_rds("data/pop_data.rds") %>% 
  subset(GEOID %in% cnty)

# Join the data to the names list and clean.

local_pop <- pop_data %>% 
  left_join(cv_names, by = "GEOID") %>% 
  select(year, county, counttype, value) %>% 
  group_by(county) %>% 
  mutate(pchg = (value - first(value))/first(value)) %>% 
  filter(year == 2020) %>% 
  select(county, value, pchg)


local_pop_plot <- ggplot(local_pop,
       aes(x = reorder(county, -value),
           y = value,
           fill = county)) +
  geom_col() +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "2020 Census population",
       caption = "**Source:** U.S. Census Bureau, Decennial Census.") +
  scale_y_continuous(labels = label_comma(), limits = c(0, 100000))

if (knitr::is_html_output()) {
  
  girage(ggobj = local_pop_plot,
         height_svg = 4)
  
} else {
  
  local_pop_plot
  
}


```

```{r}

local_growth_plot <- ggplot(local_pop,
       aes(x = reorder(county, -pchg),
           y = pchg,
           fill = county)) +
  geom_col() +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = percent_format(), limits = c(-.05, .2)) +
  labs(title = "Population growth",
       subtitle = "Percent change in population from 2010",
       caption = "**Source:** U.S. Census Bureau, Population Estimates Program.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_growth_plot,
         height_svg = 4)
  
} else {
  
  local_growth_plot
  
}


```

```{r}
#| label: fig-local-chg

local_change <- read_rds("data/comp_data.rds") %>% 
  filter(NAMELSAD != "Lynchburg city")

# Create group bar chart showing population change by locality.

local_change_plot <- ggplot(local_change,
       aes(x = year, 
           y = value,
           fill = component)) +
  geom_col(position = "stack") +
  facet_wrap(~NAMELSAD, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(size = 6, angle = 90),
        legend.position = "bottom") +
  labs(title = "Components of population change",
       caption = "**Source:** U.S. Census Bureau, Population Estimates Program.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_change_plot,
         height_svg = 4)
  
} else {
  
  local_change_plot
  
}


```

```{r}
#| label: fig-projections


projections <- read_csv("data/uva_proj.csv") %>% 
  subset(FIPS %in% cnty) %>% 
  mutate(Difference = as.numeric(Difference)) 

local_proj_plot <- ggplot(projections,
       aes(x = reorder(NAME, -Difference),
           y = Difference,
           fill = NAME)) +
  geom_col() +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Projected growth",
       subtitle = "Change from 2030 to 2050",
       caption = "University of Virginia, Weldon Cooper Center for Public Service.") +
  scale_y_continuous(labels = label_comma())

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_proj_plot,
         height_svg = 4)
  
} else {
  
  local_proj_plot 
  
}
  
```

## Household trends

```{r}
#| label: tenure

b25003_data <- read_rds("data/b25003_data.rds") %>% 
  filter(NAME != "Lynchburg city")

local_tenure <- b25003_data %>%
  select(county = NAME, year, tenure, estimate) %>% 
  group_by(county, year, tenure) %>% 
  summarise(estimate = sum(estimate)) %>% 
  group_by(county, tenure) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) %>% 
  mutate(year = as.character(year))


local_tenure_plot <- ggplot(local_tenure,
       aes(x = year,
           y = run_diff,
           fill = tenure,
           data_id = run_diff,
           tooltip )) +
  geom_col(position = "dodge") +
  facet_wrap(~county, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, size = 8),
        legend.position = "bottom") +
  labs(title = "Cumulative change in households by tenure",
       subtitle = "From 2010 to 2021",
       caption = "U.S. Census Bureau, ACS 5-Year Estimates, Table B25003.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_tenure_plot,
         height_svg = 4)
  
} else {
  
  local_tenure_plot
  
}
  

```

```{r}
#| label: age

b25007_data <- read_rds("data/b25007_data.rds") %>% 
  filter(NAME != "Lynchburg city")
  
local_age <- b25007_data %>% 
  group_by(county = NAME, year, tenure, age) %>% 
  summarise(estimate = sum(estimate)) %>% 
  filter(year == 2010 | year == 2021) %>% 
  group_by(county, year, tenure) %>% 
  mutate(pct = estimate/sum(estimate)) %>% 
  mutate(year = as.character(year))

local_age_plot <- ggplot(local_age,
       aes(x = year, 
           y = pct,
           fill = age,
           data_id = pct,
           tooltip = percent_format()(pct)))  +
  geom_col(position = "stack") +
  facet_grid(tenure~county) +
  geom_col_interactive(size = 2) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = percent_format()) +
  theme(legend.position = "bottom") +
  labs(title = "Percent of households by age and tenure",
       subtitle = "In 2010 and 2021",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25007.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_age_plot,
         height_svg = 4)
  
} else {
  
  local_age_plot
}

```

```{r}
#| label: type

b11001_data <- read_rds("data/b11001_data.rds") %>% 
  filter(NAME != "Lynchburg city")


```

```{r}
#| label: size

b25009_data <- read_rds("data/b25009_data.rds") %>% 
  filter(NAME != "Lynchburg city")

local_size <- b25009_data %>% 
  select(county = NAME, year, tenure, size, estimate) %>% 
  group_by(county, year, tenure, size) %>% 
  summarise(estimate = sum(estimate))  %>%
  group_by(county, tenure, size) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) %>% 
  mutate(year = as.character(year)) %>% 
  filter(year == "2021")


local_size_plot <- ggplot(local_size,
       aes(x = county,
           y = run_diff,
           fill = tenure,
           data_id = run_diff,
           tooltip )) +
  geom_col(position = "dodge") +
  facet_wrap(~size, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, size = 8)) +
  labs(title = "Cumulative change in households by size and tenure",
       subtitle = "From 2010 to 2021",
       caption = "U.S. Census Bureau, ACS 5-Year Estimates, Table B25009.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_size_plot,
         height_svg = 4)
  
} else {
  
  local_size_plot
}
  


```

```{r}
#| label: senior

b09020_data <- read_rds("data/b09020_data.rds") %>% 
  filter(NAME != "Lynchburg city")

local_senior <- b09020_data %>% 
  select(county = NAME, year, family, relationship, estimate) %>% 
  group_by(county, year, family, relationship) %>% 
  summarise(estimate = sum(estimate)) %>% 
  pivot_wider(names_from = year,
              values_from = estimate) %>% 
  transform(change = `2021` - `2012`) %>% 
  select(county, family, relationship, change) %>% 
  mutate(family = str_replace(family, "Group quarters", " "),
         family = fct_relevel(family, "Family", "Nonfamily", " ")) %>% 
  filter(relationship == "Living alone")

local_senior_plot <- ggplot(local_senior,
                            aes(x = county,
                                y = change,
                                fill = county,
                                data_id = change,
                                tooltip = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus"))) +
  geom_col(position = "dodge") +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")) + 
   scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(title = "Change in senior population living alone",
       subtitle = "From 2012 to 2021",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_senior_plot,
         height_svg = 4)
  
} else {
  
  local_senior_plot

}
  

```

```{r}
#| label: subfam

b11013_data <- read_rds("data/b11013_data.rds") %>% 
  filter(NAME != "Lynchburg city")

local_sub <- b11013_data %>% 
  select(county = NAME, year, subfamily, children, estimate) %>% 
  mutate(subfamily = case_when(
    subfamily == "Single father" ~ "Single parent",
    subfamily == "Single mother" ~ "Single parent",
    TRUE ~ subfamily
  )) %>% 
  group_by(county, year, subfamily, children) %>% 
  summarise(estimate = sum(estimate)) %>% 
  group_by(county, subfamily, children) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) %>% 
  filter(year == 2021) 

local_sub_plot <- ggplot(local_sub,
                         aes(x = county,
                             y = run_diff,
                             fill = county,
                             data_id = run_diff,
                             tooltip = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")(run_diff))) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  facet_wrap(children~subfamily) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus"))
  theme(legend.position = "right",
        axis.text.x = element_blank()) +
  labs(title = "Change in families living with others",
       subtitle = "From 2010 to 2021",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B11013.")


if (knitr::is_html_output()) {
  
  girafe(ggobj = local_sub_plot,
         height_svg = 4)
  
} else {
  
  local_sub_plot
  
}


```

```{r}
#| label: achild

b09021_data <- read_rds("data/b09021_data.rds") %>% 
  filter(NAME != "Lynchburg city")

local_achild <- b09021_data %>% 
  select(county = NAME, year, arrangement, estimate) %>% 
  filter(arrangement == "Child of householder") %>% 
  group_by(county, arrangement) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) 


local_achild_plot <- ggplot(local_achild,
                            aes(x = year,
                                y = run_diff,
                                fill = county,
                                data_id = run_diff,
                                tooltip = label_number(big.mark = ",", style_negative = "minus", style_positive = "plus")(run_diff))) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")) +
  theme_hfv() +
  scale_fill_hfv() +
  facet_wrap(~county, nrow = 1) +
  labs(title = "Cumulative change in adult children living with parents",
       subtitle = "From 2015 to 2021",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B09021.") +
  theme(legend.position = "right")


if (knitr::is_html_output()) {
  
  girafe(local_achild_plot,
         height_svg = 4)
  
} else {
  
  local_achild_plot
}




```


## Economic trends

```{r}
#| label: inc-dist

b25118_data <- read_rds("data/b25118_data.rds") %>% 
  filter(NAME != "Lynchburg city")

local_dist <- b25118_data %>% 
  select(county = NAME, year, tenure, income, estimate) %>% 
  group_by(county, year, tenure, income) %>% 
  summarise(estimate = sum(estimate)) %>% 
  filter(year == 2021)


level_order <- c("Less than $15,000", "$15,000 to $24,999", "$25,000 to $49,999", "$50,000 to $74,999", "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more")


local_dist_plot <- ggplot(local_dist,
       aes(x = factor(income, level = level_order),
           y = estimate,
           fill = tenure,
           data_id = estimate,
           tooltip = label_comma()(estimate))) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  facet_grid(tenure~county) +
  scale_fill_hfv() +
  theme_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = .2, size = 8)) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Income distribution",
       subtitle = "In 2021",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25118.") 

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_dist_plot,
         height_svg = 4)
} else {
  
  local_dist_plot
  
}

```

```{r}
#| label: med-inc

b25119_data <- read_rds("data/b25119_cpi.rds") %>% 
  filter(NAME != "Lynchburg city")

local_inc <- b25119_data %>% 
  select(county = NAME, year, tenure, dollars21, cdollars) %>% 
  mutate(county = str_remove_all(county, ", Virginia")) %>% 
  filter(county != "Bedford city")


local_inc_plot <- ggplot(local_inc,
                         aes(x = year,
                             y = dollars21,
                             color= tenure)) +
  geom_line() +
  geom_point_interactive(
    aes(data_id = dollars21,
        tooltip = dollar_format()(dollars21)),
    size = 1) +
  facet_wrap(~county, nrow = 1) +
  theme_hfv() +
  scale_color_hfv() +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Median household income by tenure",
       subtitle = "From 2010 to 2021 in 2021 adjusted dollars",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25119 and Bureau of Labor Statistics, Consumer Price Index Retroactive Series.") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5))

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_inc_plot,
         height_svg = 4)
} else {
  
  local_inc_plot

}


```

```{r}
#| label: race-inc

b19013_data <- read_rds("data/b19013_data.rds") %>% 
  filter(NAME != "Lynchburg city")

local_rinc <- b19013_data %>% 
  select(county = NAME, year, race, cdollars = estimate, dollars21) %>% 
  filter(year == 2021) %>% 
  drop_na() %>% 
  mutate(label = paste(county, "-", race))
  
local_rinc_plot <- ggplot(local_rinc,
       aes(x = reorder_within(county, cdollars, race),
           y = cdollars,
           fill = race,
           data_id = cdollars,
           tooltip = dollar_format()(cdollars))) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "right") +
  scale_x_reordered() +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Median household income by race and ethnicity",
       subtitle = "In 2021 inflation-adjusted dollars",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B19013B - I.")

if (knitr::is_html_output()){
  
  girafe(ggobj = local_rinc_plot,
         height_svg = 4)
  
} else {
  
  local_rinc_plot
}
  


```


## Housing stock

```{r}
#| label: structure

b25127_data <- read_rds("data/b25127_data.rds") 

local_str <- b25127_data %>% 
  select(county = NAME, year, tenure, structure, estimate) %>% 
  group_by(county, year, tenure, structure) %>% 
  summarise(estimate = sum(estimate)) %>% 
  filter(year == 2010 | year == 2021) %>% 
  group_by(year) %>% 
  transform(pct = estimate/sum(estimate)) %>% 
  mutate(county = fct_reorder(county, pct, .desc = TRUE)) %>% 
  filter(year == 2021) %>% 
  filter(county != "Lynchburg city")

level_order <- c("Single-family", "2 to 4 units", "5 to 19 units", "20 or more units", "Other")

local_str_plot <- ggplot(local_str,
       aes(x = factor(structure, level = level_order),
           y = pct,
           fill = county)) +
  geom_col(position = "stack") +
  facet_wrap(~tenure, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() + 
  scale_y_continuous(labels = percent_format()) +
  theme(legend.position = "right") + 
   scale_x_discrete(labels = function(x) str_wrap(x, width = 8)) +
  labs(title = "Distribution of the housing stock in the region",
      subtitle = "Percent of total occupied housing in 2021",
      caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25127.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_str_plot,
         height_svg = 4)
  
} else {
  
  local_str_plot
  
}


```

```{r}
#| label: mhc


mhc <- st_read("https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Mobile_Home_Parks/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson", quiet = TRUE) |> 
  filter(STATE == "VA") %>% 
  filter(str_detect(COUNTY, "AMHERST|APPOMATTOX|BEDFORD|CAMPBELL"))



```

```{r}
#| label: fig-mhc-county


cv_mhc <- mhc  %>% 
  group_by(COUNTY, SIZE) %>% 
  summarise(count = n_distinct(OBJECTID)) %>% 
  st_drop_geometry() %>% 
  mutate(COUNTY = str_to_sentence(COUNTY))

ggplot(cv_mhc,
       aes(x = reorder(COUNTY,count),
           y = count,
           fill = SIZE)) +
  geom_col(position = "stack") +
  theme_hfv() +
  scale_fill_hfv()


```

```{r}
#| label: bps

cv_cbps <- read_rds("data/cv_cbps.rds")

```

## Homeownership market

```{r}
#| label: ho

b25003_data <- read_rds("data/b25003_data.rds") 

```


## Rental market

```{r}
#| label: rent

# Utilize CoStar properties search for Multi-Family. Filter by County for Lynchburg - VA Market. Then review Analytics and pull QUARTERLY Market Asking Rent Per Unit for entire region. Do not bring in Forecast and set History to All. Combine Bedford City and Bedford County.

# Download data for current rent and inflation-adjusted rent into a single csv - name [locality_abbv]_rent.csv --- for example amherst_rent.csv or lburg_rent.csv

# Merge all data into a single csv named - local_rent.csv

# Note the download date of data here: 1/23/2023

# lburg_rent <- read_csv("data/lburg_rent.csv") %>% 
#  mutate(county = "Lynchburg city")

amherst_rent <- read_csv("data/amherst_rent.csv") %>% 
  mutate(county = "Amherst County")

appomattox_rent <- read_csv("data/appomattox_rent.csv") %>% 
  mutate(county = "Appomattox County")

bedford_rent <- read_csv("data/bedford_rent.csv") %>% 
  mutate(county = "Bedford County")

campbell_rent <- read_csv("data/campbell_rent.csv") %>% 
  mutate(county = "Campbell County")

library(zoo)

local_rent <- rbind(lburg_rent, amherst_rent, appomattox_rent, bedford_rent, campbell_rent) %>% 
  filter(!str_detect(period, "QTD")) %>% 
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) %>% 
  select(county, period, current, inf) %>% 
  pivot_longer(3:4,
               names_to = "value",
               values_to = "rent") %>% 
  mutate(value = case_when(
    value == "current" ~ "Current dollars",
    value == "inf" ~ "Inflation-adjusted dollars"
  ))

ggplot(local_rent,
       aes(x = period,
           y = rent,
           group = value,
           color = county)) +
  geom_line(aes(linetype = value)) + 
  facet_wrap(~county, nrow = 1) +
  scale_linetype_manual(values = c("dotted", "solid")) +
  theme_hfv() +
  scale_color_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Quarterly average rent",
       subtitle = "Market asking rent from 2000 through 2022",
       caption = "**Source:** CoStar Group Inc.")



```

```{r}
#| label: vacancy

# Utilize CoStar properties search for Multi-Family. Filter by County for Lynchburg - VA Market. Then review Analytics and pull QUARTERLY Vacancy Rate for entire region. Do not bring in Forecast and set History to All.

# Download data into csv - name [locality_abbv]_vacancy.csv --- for example amherst_vacancy.csv or lburg_vacancy.csv

# Note the download date of data here: 




```


```{r}
#| label: nhpd

nhpd_subsidies <- read_csv("data/nhpd_subsidies.csv") %>% 
  clean_names() %>% 
  filter(subsidy_status != "Inactive") %>% 
  filter(grepl("amherst", city, ignore.case = TRUE) |
           grepl("bedford", city, ignore.case = TRUE) |
           grepl("campbell", city, ignore.case = TRUE) |
           grepl("appomattox", city, ignore.case = TRUE)) %>% 
  select(subsidy_name, subsidy_subname, end_date, units = assisted_units, property_name, owner_type, target_population, city, longitude, latitude)


cnty_nhpd <- nhpd_subsidies %>% 
  group_by(city, subsidy_name) %>% 
  summarise(units = sum(units))

ggplot(cnty_nhpd,
       aes(x = city, 
           y = units,
           fill = subsidy_name)) +
  geom_col(position = "stack") +
  theme_hfv() +
  scale_fill_hfv()

```




