# City of Lynchburg Housing Market Analysis

The following provides an analysis of major trends impacting housing within the City of Lynchburg. 

```{r}

# THINGS TO UPDATE ----------

# Update plot styles to match previous chapters


# HOUSEHOLD

# Add ACS selected disability (2010-2021) -- maybe collapse ages into just <65 and 65+
# - Ambulatory difficulty (B18105 or C18105)
# - Independent Living difficulty (B18107 or C18107)

# ECONOMIC

# Add QCEW data -- total employment and average wage?

# Add LEHD data -- commute inflow/outflow?
# See lehd examples in sandbox repo

# Add LAUS unemployment data (low priority, but in our proposal)

# HOUSING STOCK

# Add Census BPS through March 2023 (or just through end of 2022)

# Add ACS overall vacancy rates for owner- and renter-occupied (2010-2021)

# Add ACS vacancy status/reason (B25004, 2010-2021) -- split into 3 plots?
# 1. Total units for rent and rented, not occupied
# 2. Total units for sale only or sold, not occupied
# 3. Total units for seasonal, for migrant workers, or other vacant (not split by tenure)

# Update map to style used here https://hdadvisors.github.io/rrh-framework/part-2-4.html#manufactured-home-communities

# HOMEOWNERSHIP

# Add specific home market data pending transaction-level MLS data
# - New build vs resale prices (and by bedrooms, sqft)

# Mortgage delinquency from CPFB (low priority, but in our proposal)

# Add HMDA data (low priority, but in our proposal)

# RENTAL

# Add single-family rent estimates from ACS for MSA

# Add details on public housing (map, units by community, age)

# Add more preservation details
# - More on non-LIHTC properties that could expire (i.e. everything not public housing)
# - Map of LIHTC props colored by exp year
# - Map of Section 8 props colored by exp year

# Move rental housing gap plot to next section

# AFFORDABILITY

# From proposal:
# "Current and projected affordable rental housing gaps within the city
# to determine the number of new units needed now and into the future"
# - Not sure what to do about "projected"

```

```{r}
#| label: setup

library(tidyverse)
library(tidyverse)
library(janitor)
library(ggplot2)
library(leaflet)
library(tigris)
library(sf)
library(ggiraph)
library(hdatools)
library(scales)
library(readxl)
library(ggtext)
library(tidytext)
library(lubridate)
library(zoo)

lburg <- "51680"

```

## Takeaways

- Growth in the city has been a result of international migration and natural increases.
- Once 44 percent renter households, the city is now majority renter households at 51 percent.
- There has been a loss of homeowners in the city, mainly among those households aged 25 to 64 years old.
- The number of higher income households is growing in the city, even among renters, but there is also a growing number of extremely low-income renters.
- In spite of its urban environment, housing in the city is dominated by single-family housing.
- Homeownership is declining as prices have climbed rapidly during the pandemic.
- Large amount of detached single-family rentals, especially in College Hill and Diamond Hill areas.
- Rental prices have remained generally flat among multifamily properties, but cost burden among renters has continued to rise.


## Population trends

From 2010 to 2020, Lynchburg has only grown by four percent to reach a population of 79,009. Despite this growth, the current population is an estimated decline from a 2019 estimated population of 82,168. 
```{r}
#| label: fig-pop-2020
#| fig-cap: "Total population in Lynchburg"

# Read in population data and filter for Lynchburg.

pop_data <- read_rds("data/pop_data.rds") %>% 
  subset(GEOID %in% lburg)

# Join to names data frame (just in case) then select only the fields needed.

local_pop <- pop_data %>% 
  select(year, counttype, value) 

# Read in projections data to include in visualization and clean the data.

projections <- read_csv("data/uva_proj.csv") %>% 
  subset(FIPS %in% lburg) %>% 
  select(`2030`, `2040`, `2050`) %>% 
  pivot_longer(1:3,
               names_to = "year",
               values_to = "value") %>% 
  mutate(counttype = "Projection") %>% 
  select(year, counttype, value)

# Combine population estimates and projections into single data frame.

lburg_pop <- rbind(local_pop, projections)

# Visualize as colored bar chart. Showing different colors for count type.

lburg_po <- lburg_pop |> 
  filter(counttype != "Projection")
  
local_pop_plot <- ggplot(lburg_po,
       aes(x = year,
           y = value,
           fill = counttype,
           data_id = value,
           tooltip = label_number(big.mark = ",")(value))) +
  geom_col() +
  geom_col_interactive() +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Total population in Lynchburg",
       caption = "**Source:** U.S. Census Bureau, Decennial Census and Population Estimates Program and UVA, Weldon Cooper Center.") +
  scale_y_continuous(labels = label_comma(), limits = c(0, 100000)) +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 90, vjust = 0.5))

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_pop_plot,
         height_svg = 4)
  
} else {
  
  local_pop_plot
  
}


```
The growth in population over the last decade in Lynchburg has been a result of international migration and natural increases. This is in stark contrast to the rest of the region where growth has been driven by domestic migration.

```{r}
#| label: fig-local-chg
#| fig-cap: "Components of population change in Lynchburg"


# Read in components of population change data and filter for Lynchburg.

local_change <- read_rds("data/comp_data.rds") %>% 
  filter(NAMELSAD == "Lynchburg city")

# Create group bar chart showing population change by locality.

local_change_plot <- ggplot(local_change,
       aes(x = year, 
           y = value,
           fill = component)) +
  geom_col(position = "stack") +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom") +
  labs(title = "Components of population change in Lynchburg",
       caption = "**Source:** U.S. Census Bureau, Population Estimates Program.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_change_plot,
         height_svg = 4)
  
} else {
  
  local_change_plot
  
}

```
Despite the decrease in population from 2019 to 2020, the city is expected to see continued growth over the next few decades. By 2050, the city is projected to have a population of 93,708, a 19 percent increase over 30 years.

```{r}
#| label: fig-pop-proj
#| fig-cap: "Projected population in Lynchburg"

lburg_p <- lburg_pop |> 
  filter(counttype == "Projection" | counttype == "Census population") |> 
  mutate(year == as.character(year))

local_pop_plot <- ggplot(lburg_p,
       aes(x = year,
           y = value,
           fill = counttype,
           data_id = value,
           tooltip = label_number(big.mark = ",")(value))) +
  geom_col() +
  geom_col_interactive() +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Projected population in Lynchburg",
       caption = "**Source:** U.S. Census Bureau, Decennial Census and Population Estimates Program and UVA, Weldon Cooper Center.") +
  scale_y_continuous(labels = label_comma(), limits = c(0, 100000)) +
  theme(legend.position = "right")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_pop_plot,
         height_svg = 4)
  
} else {
  
  local_pop_plot
  
}



```

## Household trends

Lynchburg has steadily lost homeowners, while gaining renters. By 2021, the city had gained a total of 2,563 renters since 2010, while subsequently losing 1,499 homeowners. These changes have left the balance of renter and homeowner households in slight favor of renters, as renter households make up 51 percent of households in the city.This is shift from 2010 when renter households made up 44 percent of all households in the city.
 
```{r}
#| label: fig-tenure
#| fig-cap: "Change in households by tenure in Lynchburg"

# Read in household tenure data and filter for Lynchburg.

b25003_data <- read_rds("data/b25003_data.rds") %>% 
  filter(NAME == "Lynchburg city")

# Clean data and calculate cumulative change in households by tenure.

local_tenure <- b25003_data %>%
  select(county = NAME, year, tenure, estimate) %>% 
  group_by(county, year, tenure) %>% 
  summarise(estimate = sum(estimate)) %>% 
  group_by(county, tenure) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) %>% 
  mutate(year = as.character(year))

# Create bar chart visualization that shows the cumulative change from 2010 to 2021.

local_tenure_plot <- ggplot(local_tenure,
       aes(x = year,
           y = run_diff,
           fill = tenure,
           data_id = run_diff,
           tooltip = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")(run_diff))) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge") +
  facet_wrap(~county, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, size = 8),
        legend.position = "bottom") +
  labs(title = "Cumulative change in households by tenure in Lynchburg",
       subtitle = "From 2010 to 2021",
       caption = "U.S. Census Bureau, ACS 5-Year Estimates, Table B25003.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_tenure_plot,
         height_svg = 4)
  
} else {
  
  local_tenure_plot
  
}
  

```
Breaking down the changes in tenure by age shows that more and more households of all ages are now renting in the city. The largest increases were among renter households under 45 years old, an obvious result of the many higher education institutions in the city.

Losses were mainly among 25 to 64 year old homeowners. By 2021, the city had lost 603 25 to 44 year old homeowner households and 909 45 to 64 year old homeowner households. 

```{r}
#| label: fig-age
#| fig-cap: "Cumulative change in households by tenure and age in Lynchburg"

b25007_data <- read_rds("data/b25007_data.rds") %>% 
  filter(NAME == "Lynchburg city")

local_age <- b25007_data |> 
  group_by(county = NAME, year, tenure, age) |> 
  summarise(estimate = sum(estimate)) |> 
  group_by(tenure, age) |> 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) |> 
  mutate(run_diff = cumsum(diff)) |>  
  filter(run_diff != 0) |> 
  mutate(year = as.character(year))


# local_age <- b25007_data %>% 
#   group_by(county = NAME, year, tenure, age) %>% 
#   summarise(estimate = sum(estimate)) %>% 
#   filter(year == 2010 | year == 2021) %>% 
#   group_by(county, year, tenure) %>% 
#   mutate(pct = estimate/sum(estimate)) %>% 
#   mutate(year = as.character(year))

local_age_plot <- ggplot(local_age,
       aes(x = year, 
           y = run_diff,
           fill = age,
           data_id = run_diff,
           tooltip = run_diff))  +
  geom_col(position = "dodge") +
  facet_grid(~tenure) +
  geom_col_interactive(position = "dodge", size = 2) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = number_format(big.mark = ",", style_positive = "plus")) +
  theme(legend.position = "right") +
  labs(title = "Cumulative change in households by age and tenure in Lynchburg",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25007.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_age_plot,
         height_svg = 4)
  
} else {
  
  local_age_plot
}

```

```{r}
#| label: type
#| eval: false

b11001_data <- read_rds("data/b11001_data.rds") %>% 
  filter(NAME != "Lynchburg city")


```
Among those growing number of renters, most of those households have been two-person households, an addition of 2,011 from 2010 to 2021. Households of three people or less have been declining among homeowners, but four or more person homeowner households surprisingly grew by 409 during this time.
```{r}
#| label: figsize
#| fig-cap: "Change in households by size and tenure in Lynchburg"

b25009_data <- read_rds("data/b25009_data.rds") %>% 
  filter(NAME == "Lynchburg city")

local_size <- b25009_data %>% 
  select(county = NAME, year, tenure, size, estimate) %>% 
  group_by(county, year, tenure, size) %>% 
  summarise(estimate = sum(estimate))  %>%
  group_by(county, tenure, size) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) %>% 
  mutate(year = as.character(year))


local_size_plot <- ggplot(local_size,
       aes(x = year,
           y = run_diff,
           fill = tenure,
           data_id = run_diff,
           tooltip = number_format(big.mark = ",")(run_diff))) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  facet_grid(tenure~size) +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, size = 8)) +
  labs(title = "Cumulative change in households by size and tenure in Lynchburg",
       subtitle = "From 2010 to 2021",
       caption = "U.S. Census Bureau, ACS 5-Year Estimates, Table B25009.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_size_plot,
         height_svg = 4)
  
} else {
  
  local_size_plot
}
  


```
The city has experienced a decreasing number of seniors living alone or with nonrelatives in recent years (-575). This is another contrast to the surrounding counties, where more and more seniors are living alone. Whether this is the result of Lynchburg seniors moving to the counties is not easily discerned, but regardless seniors in the city may find it more difficult to find housing that meets their needs for aging-in-place. Seniors in the city are rather increasingly the head of household (+604) or living with a spouse (+473).

```{r}
#| label: fig-senior
#| fig-cap: "Change in senior population in Lynchburg"

b09020_data <- read_rds("data/b09020_data.rds") %>% 
  filter(NAME == "Lynchburg city")

local_senior <- b09020_data %>% 
  select(county = NAME, year, family, relationship, estimate) %>% 
  group_by(county, year, family, relationship) %>% 
  summarise(estimate = sum(estimate)) %>% 
  pivot_wider(names_from = year,
              values_from = estimate) %>% 
  transform(change = `2021` - `2012`) %>% 
  select(county, family, relationship, change) %>% 
  mutate(family = str_replace(family, "Group quarters", " "),
         family = fct_relevel(family, "Family", "Nonfamily", " "))


local_senior_plot <- ggplot(local_senior,
       aes(y = reorder_within(relationship, change, family),
           x = change,
           fill = family,
           data_id = change,
           tooltip = change)) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge", size = 2) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_reordered() +
  scale_x_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")) +
  facet_grid(rows = vars(family), scales = "free_y", space = "free", switch = "y") +
  #scale_y_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(title = "Change in senior population by living arrangement in Lynchburg",
       subtitle="From 2012 to 2021",
       fill="Tenure",
       caption = "**Source:** U.S. Census Bureau, American Community Survey, Table B09020.")  +
  theme(strip.placement = "outside",
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(color = "#e2e4e3",
                                          size = 0.05)) +
  add_zero_line(axis = "x")
                              

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_senior_plot,
         height_svg = 4)
  
} else {
  
  local_senior_plot

}
  

```
The number of single parents living with other families is increasing in the city. From 2010 to 2021, there has been an increase of 330 single parents having to live with others.
```{r}
#| label: fig-subfam
#| fig-cap: "Cumulative change in families living with others in Lynchburg"

b11013_data <- read_rds("data/b11013_data.rds") %>% 
  filter(NAME == "Lynchburg city")

local_sub <- b11013_data %>% 
  select(county = NAME, year, subfamily, children, estimate) %>% 
  mutate(subfamily = case_when(
    subfamily == "Single father" ~ "Single parent",
    subfamily == "Single mother" ~ "Single parent",
    TRUE ~ subfamily
  )) %>% 
  group_by(county, year, subfamily, children) %>% 
  summarise(estimate = sum(estimate)) %>% 
  group_by(county, subfamily, children) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) |> 
  mutate(year = as.character(year))

local_sub_plot <- ggplot(local_sub,
                         aes(x = year,
                             y = run_diff,
                             fill = subfamily,
                             data_id = run_diff,
                             tooltip = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")(run_diff))) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  facet_wrap(~children) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  labs(title = "Cumulative change in families living with others in Lynchburg",
       subtitle = "From 2010 to 2021",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B11013.")


if (knitr::is_html_output()) {
  
  girafe(ggobj = local_sub_plot,
         height_svg = 4)
  
} else {
  
  local_sub_plot
  
}


```
Although the city was seeing a significant increase in adult (18 to 34 years old) children living with parents between 2016 and 2020, the overall increase was only 558 by 2021, when there were 4,730. These living arrangements, as well as families living with others, are often an indicator of affordability challenges in communities. The increase in 2020 may have likely been a result of pandemic impacts, and the subsequent decrease by 2021 could have been the result of loosening pandemic restrictions.
```{r}
#| label: fig-achild
#| fig-cap: "Cumulative change in adult children living with parents in Lynchburg"

b09021_data <- read_rds("data/b09021_data.rds") %>% 
  filter(NAME == "Lynchburg city")

local_achild <- b09021_data %>% 
  select(county = NAME, year, arrangement, estimate) %>% 
  filter(arrangement == "Child of householder") %>% 
  group_by(county, arrangement) %>% 
  mutate(diff = estimate - lag(estimate),
         diff = replace_na(diff, 0)) %>% 
  mutate(run_diff = cumsum(diff)) %>%  
  filter(run_diff != 0) %>% 
  mutate(year = as.character(year))


local_achild_plot <- ggplot(local_achild,
                            aes(x = year,
                                y = run_diff,
                                fill = run_diff,
                                data_id = run_diff,
                                tooltip = label_number(big.mark = ",", style_negative = "minus", style_positive = "plus")(run_diff))) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")) +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Cumulative change in adult children living with parents in Lynchburg",
       subtitle = "From 2015 to 2021",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B09021.") +
  scale_fill_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E"))


if (knitr::is_html_output()) {
  
  girafe(ggobj = local_achild_plot,
         height_svg = 4)
  
} else {
  
  local_achild_plot
}


```

## Economic trends

Income distribution in the city has shifted from 2010 to 2021. Lower income homeowner households have decreased over a decade, resulting in fewer homeowners making less than \$25,000. And while there has been an increase in lower income renter households, there has also been an increase in renter households making \$75,000 or more.

```{r}
#| label: fig-inc-dist
#| fig-cap: "Income distribution by tenure in Lynchburg"

b25118_data <- read_rds("data/b25118_data.rds") %>% 
  filter(NAME == "Lynchburg city")

local_dist <- b25118_data %>% 
  select(county = NAME, year, tenure, income, estimate) %>% 
  group_by(county, year, tenure, income) %>% 
  summarise(estimate = sum(estimate)) %>% 
  filter(year == 2010 | year == 2021) %>% 
  mutate(year = as.character(year))


level_order <- c("Less than $15,000", "$15,000 to $24,999", "$25,000 to $49,999", "$50,000 to $74,999", "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more")


local_dist_plot <- ggplot(local_dist,
       aes(x = factor(income, level = level_order),
           y = estimate,
           fill = tenure,
           data_id = estimate,
           tooltip = label_comma()(estimate))) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  facet_grid(tenure~year) +
  scale_fill_hfv() +
  theme_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = .2, size = 8)) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Income distribution by tenure in Lynchburg",
       subtitle = "In 2010 and 2021",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25118.") 

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_dist_plot,
         height_svg = 4)
} else {
  
  local_dist_plot
  
}

```
Median household income emphasizes these shifts. Homeowner median household income has increased by 15 percent from 2010 to 2021, while renter income increased by 25 percent. The growth in the median renter income was the result of higher income renters coming into the city, but still the typical renter in the city only makes about half (\$38,229) of what a homeowner makes (\$75,093).
```{r}
#| label: fig-med-inc
#| fig-cap: "Median household income by tenure in Lynchburg"

b25119_data <- read_rds("data/b25119_cpi.rds") %>% 
  mutate(NAME = str_remove_all(NAME, ", Virginia")) %>% 
  filter(NAME == "Lynchburg city")

local_inc <- b25119_data %>% 
  select(county = NAME, year, tenure, dollars21, cdollars)



local_inc_plot <- ggplot(local_inc,
                         aes(x = year,
                             y = dollars21,
                             color= tenure)) +
  geom_line() +
  geom_point_interactive(
    aes(data_id = dollars21,
        tooltip = dollar_format()(dollars21)),
    size = 1) +
  theme_hfv() +
  scale_color_hfv() +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Median household income by tenure in Lynchburg",
       subtitle = "In 2021 adjusted dollars",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25119 and Bureau of Labor Statistics, Consumer Price Index Retroactive Series.") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5))

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_inc_plot,
         height_svg = 4)
} else {
  
  local_inc_plot

}


```
In the city, Hispanic households have seen major increases in median household income since 2010. As of 2021, Hispanic households have the highest median income at \$72,680. White, non-Hispanic households follow at $62,857, a slight increase of about \$9,000 from 2010. 

Black households in the city, however, had consistently low median household incomes. From 2010 to 2021, the Black median household income had only increased by 14 percent, while their white counterparts saw an 18 percent increase, furthering the Black-white income gap.

:::{.callout-warning}

Note about margins of error. Ways people identify may have impacted this major increase.

:::
```{r}
#| label: fig-race-inc
#| fig-cap: "Median household income by race and ethnicity in Lynchburg"

b19013_data <- read_rds("data/b19013_data.rds") %>% 
  filter(NAME == "Lynchburg city")

local_rinc <- b19013_data %>% 
  select(county = NAME, year, race, cdollars = estimate, dollars21) %>% 
  filter(year == 2021 | year == 2010) %>% 
  drop_na()
  
local_rinc_plot <- ggplot(local_rinc,
       aes(x = reorder_within(race, dollars21, year),
           y = dollars21,
           fill = race,
           data_id = dollars21,
           tooltip = dollar_format()(dollars21))) +
  geom_col(stat = "identity", position = "dodge") +
  geom_col_interactive() +
  theme_hfv() +
  scale_fill_hfv() + 
  scale_x_reordered() +
  facet_wrap(~year, scales = "free_x") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "right") +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Median household income by race and ethnicity in Lynchburg",
       subtitle = "In 2021 inflation-adjusted dollars",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B19013B - I.")

if (knitr::is_html_output()){
  
  girafe(ggobj = local_rinc_plot,
         height_svg = 4)
  
} else {
  
  local_rinc_plot
}
  

```


## Housing stock

Lynchburg contributes to 28 percent of the region's total housing stock in 2021. Most of that contribution is composed of single-family housing that is in the homeownership stock. Despite it's urban environment, the homeownership stock in the city is not diverse and very few duplexes and other smaller housing types are not utilized for homeownership.

Rather, diverse housing is largely used for rental housing. Most of the rental housing stock in the city is made up of single-family and medium-scale multifamily (5 to 19 units). 

```{r}
#| label: fig-structure
#| fig-cap: "Distribution of housing stock in Lynchburg in relation to the region"

b25127_data <- read_rds("data/b25127_data.rds") 

local_str <- b25127_data %>% 
  select(county = NAME, year, tenure, structure, estimate) %>% 
  group_by(year, county, tenure, structure) %>% 
  summarise(estimate = sum(estimate)) %>% 
  filter(year == 2010 | year == 2021) %>% 
  group_by(year) %>% 
  mutate(pct = estimate/sum(estimate)) %>% 
  mutate(county = fct_reorder(county, pct, .desc = TRUE)) %>% 
  filter(year == 2021) %>% 
  filter(county == "Lynchburg city")

level_order <- c("Single-family", "2 to 4 units", "5 to 19 units", "20 or more units", "Other")

local_str_plot <- ggplot(local_str,
       aes(x = factor(structure, level = level_order),
           y = pct,
           fill = tenure,
           data_id = pct,
           tooltip = percent_format()(pct))) +
  geom_col(position = "stack") +
  geom_col_interactive() +
  facet_wrap(~tenure, nrow = 1) +
  theme_hfv() +
  scale_fill_hfv() + 
  scale_y_continuous(labels = percent_format()) +
  theme(legend.position = "right",
        strip.text = element_blank()) + 
   scale_x_discrete(labels = function(x) str_wrap(x, width = 8)) +
  labs(title = "Distribution of housing stock in Lynchburg in relation to the region",
      subtitle = "Percent of total occupied housing in 2021",
      caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25127.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_str_plot,
         height_svg = 4)
  
} else {
  
  local_str_plot
  
}


```
Manufactured home communities exist within the city limits, mainly on the outskirts along major thoroughfares. Six of the eleven communities are located along or in proximity to Lakeside Drive in the western end of the city.
```{r}
#| label: fig-mhc
#| fig-cap: "Map of manufactured home communities in Lynchburg"


mhc <- st_read("https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Mobile_Home_Parks/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson", quiet = TRUE) |> 
  filter(STATE == "VA") %>% 
  filter(str_detect(COUNTY, "LYNCHBURG"))

leaflet(mhc) %>% 
  addTiles() %>% 
  addCircleMarkers()



```
The majority of communities are small, consisting of fewer than 50 homes. Although not as prevalent as in the counties, manufactured home communities also provide an affordable housing option within the city.
```{r}
#| label: fig-mhc-county
#| fig-cap: "Manufactured home communities by size in Lynchburg"


lburg_mhc <- mhc%>% 
  group_by(SIZE) %>% 
  summarise(count = n_distinct(OBJECTID)) %>% 
  st_drop_geometry()

lburg_mhc_plot <- ggplot(lburg_mhc,
       aes(x = SIZE,
           y = count,
           fill = SIZE,
           data_id = count,
           tooltip = count)) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(breaks = seq(0,15, by = 2), limits = c(0,15)) +
  labs(title = "Manufactured home communities by size in Lynchburg")


if (knitr::is_html_output()) {
  
  girafe(ggobj = lburg_mhc_plot,
         height_svg = 4)
  
} else {
  
  lburg_mhc_plot
  
}


```
Residential building permits display the dominance of single-family construction within the city for much of the past two decades. Recovery from the Great Recession has still yet to come to the city, but the construction of five or more unit buildings in the city has continued to surpass all other types in recent years. Although the city saw the construction of duplexes and triplexes early on in the 2000s, there have been little to no construction of this type of housing in the city.
```{r}
#| label: fig-bps
#| fig-cap: "Residential building permits by unit type in Lynchburg"

lburg_cbps <- read_rds("data/cv_cbps.rds") %>% 
  filter(NAMELSAD == "Lynchburg city") 

lburg_cbps_plot <- ggplot(lburg_cbps,
       aes(x = year,
           y = units, 
           fill = type, 
           data_id = units,
           tooltip = label_comma()(units))) +
  geom_col(position = "dodge") + 
  geom_col_interactive() +
  facet_grid(~type) +
  theme_hfv() +
  scale_fill_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8)) +
  scale_x_continuous(breaks = seq(from = 2000, to = 2022, by = 2.0)) +
  labs(title = "Residential building permits by unit type in Lynchburg",
       subtitle = "From 2000 through 2022",
       caption = "**Source:** U.S. Census Bureau, Building Permits Survey.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = lburg_cbps_plot,
         height_svg = 4)
  
} else {
  
  lburg_cbps_plot
}

```

## Homeownership market

As the number of renters has been increasing in the city, the homeownership rate has continued to drop. From 56 percent in 2010 to 49 percent in 2021, the city is now majority-renter.


```{r}
#| label: fig-ho
#| fig-cap: "Homeownership rate in Lynchburg"

# Read in saved tenure data.

b25003_data <- read_rds("data/b25003_data.rds") 

# Clean the data and calculate homeownership rate.

ho_rate <- b25003_data %>%
  select(county = NAME, year, tenure, estimate) %>% # Change "NAME" field to "county".
  filter(county == "Lynchburg city") %>% # Filter for Lynchburg only.
  group_by(year) %>% # Calculate percent of total by year to get homeownership rate.
  mutate(pct = estimate/sum(estimate)) %>%  
  filter(tenure == "Homeowner") # Retain only homeowners to get the homeownership rate.

cv_ho_plot <- ggplot(ho_rate,
       aes(x = year, 
           y = pct,
           group = 1,
           data_id = pct,
           color = "#011E41",
           tooltip = percent_format()(pct))) +
  geom_line(size = 0.5) +
  geom_point_interactive(
    aes(data_id = pct,
        tooltip = percent_format()(pct)),
    size = 1) +
  theme_hfv() +
  scale_color_hfv() +
  scale_y_continuous(labels = percent_format()) +
  scale_x_continuous(breaks = seq(from = 2010, to = 2021, by = 1)) +
  labs(title = "Homeownership rate in Lynchburg",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25003.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_ho_plot,
         height_svg = 4)
} else {
  
  cv_ho_plot
}


```

As the homeownership rate has declined, home prices have been on the rise. The pandemic had a profound impact on the homeownership market in the city. The median home price increased by 10 percent from 2019 to 2020 and by 2022, the price had reached \$240,000, an overall 22 percent increase from 2019. Previously, the city was seeing increases closer to one percent before the pandemic. 
```{r}
#| label: fig-lburg-sales
#| fig-cap: "Lynchburg median home sales price"

lburg_price <- read_excel("data/lburg_mls_data.xlsx", sheet = "Median Sales Price 2008 - 2022") %>% 
  clean_names() %>% 
  filter(property_type == "RESI") %>% 
  select(year = year_sd, medprice = median_selling_price)


library(fredr)
library(lubridate)

cpi_ls <- fredr(
  series_id = "CUUR0000SA0L2"
) %>% 
  select(date, value) %>% 
  mutate(date = as.Date(date)) %>% 
  mutate(value = as.numeric(value)) %>% 
  mutate(year = year(date)) %>% 
  group_by(year) %>% 
  summarise(cpi = mean(value))

price_adj <- lburg_price %>% 
  left_join(cpi_ls, by = "year") %>% 
  mutate(price_adj = (271.6895/cpi)*medprice) %>% 
  select(year, price_adj) %>% 
  mutate(price_adj = round(as.numeric(price_adj))) %>% 
  mutate(tooltip = c(paste0("Year: ", year, "\n Median sales price: $", price_adj)))

price_adj_plot <- ggplot(price_adj,
       aes(x = year,
           y = price_adj,
           color = price_adj)) +
  geom_line(size = 1) +
  geom_point_interactive(aes(
    size = .8,
    data_id = tooltip,
    tooltip = tooltip
  )) +
  theme_hfv() +
  scale_color_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Lynchburg median home sales price",
       caption = "**Source:** Lynchburg REALTORS Multiple Listing Service (MLS) and <br>U.S. Bureau of Labor Statistics, Consumer Price Index for All Urban Consumers: <br>All Items Less Shelter in U.S. City Average [CUUR0000SA0L2].")


if (knitr::is_html_output()) {
  
  girafe(ggobj = price_adj_plot,
         height_svg = 4)
  
} else {
  
  price_adj_plot
}

```
```{r}
#| label: lburg-mprice
#| eval: false

lburg_mprice <- read_excel("data/lburg_mls_data.xlsx", sheet = "Median Sales Price By Month") %>% 
  clean_names() %>% 
  filter(property_type == "RESI") %>% 
  select(year = year_sd, month = month_sd, medprice = median_selling_price) %>% 
  mutate(date = as.yearmon(paste(year, month), "%Y %m")) %>% 
  mutate(date = as.Date(date)) %>% 
  select(date, medprice)


ggplot(lburg_mprice,
       aes(x = date,
           y = medprice)) +
  geom_line()
  

```

Closed home sales increased slightly during the pandemic reaching a high of 191 homes sold in the June 2021. Despite the high demand produced by record low interest rates, the number of homes for sale in the city did not increase significantly. 

```{r}
#| label: fig-lburg-sold
#| fig-cap: "Lynchburg closed home sales"

lburg_sold <- read_excel("data/lburg_mls_data.xlsx", sheet = "Num Sold Listings Per Month") %>% 
  clean_names() %>% 
  filter(property_type == "RESI") %>% 
  select(year = year_sd, month = month_sd, sold = number_sold_listings) %>% 
  mutate(date = as.yearmon(paste(year, month), "%Y %m")) %>% 
  mutate(date = as.Date(date)) %>% 
  select(date, sold) %>% 
  mutate(tooltip = c(paste0("Date: ", date, "\n Homes sold: ", sold)))


lburg_sold_plot <- ggplot(lburg_sold,
       aes(x = date,
           y = sold, 
           fill = sold,
           data_id = tooltip,
           tooltip = tooltip)) +
  geom_col() +
  geom_col_interactive() +
  theme_hfv() +
  scale_fill_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) +
  labs(title = "Lynchburg closed home sales",
       caption = "**Source:** Lynchburg REALTORS Multiple Listing Service (MLS).")


if (knitr::is_html_output()) {
  
  girafe(ggobj = lburg_sold_plot,
         height_svg = 4)
  
} else {
  
  lburg_sold_plot
}



```

The demand in the city is exemplified by the average days on market, which was already trending downward ahead of the pandemic. But ever since the start of the pandemic, the average days on market has generally not exceeded more than 30 days.

```{r}
#| label: fig-dom
#| fig-cap: "Lynchburg average days on market"


lburg_dom <- read_excel("data/lburg_mls_data.xlsx", sheet = "Average DOM") %>% 
  clean_names() %>% 
  filter(property_type == "RESI")%>% 
  select(year = year_sd, month = month_sd, dom = average_dom) %>% 
  mutate(date = as.yearmon(paste(year, month), "%Y %m")) %>% 
  mutate(date = as.Date(date)) %>% 
  select(date, dom)


lburg_dom_plot <- ggplot(lburg_dom,
       aes(x = date,
           y = dom,
           fill = dom,
           data_id = dom,
           tooltip = dom)) +
  geom_col() + 
  geom_col_interactive() +
  theme_hfv() +
  scale_fill_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) +
  labs(title = "Lynchburg average days on market",
       caption = "**Source:** Lynchburg REALTORS Multiple Listing Service (MLS).")

if (knitr::is_html_output()) {
  
  girafe(ggobj = lburg_dom_plot,
         height_svg = 4)
  
} else {
  
  lburg_dom_plot
  
}


```
From 2016 through 2022, homes sales in the city have been concentrated in the Perrymont and Fort Hill neighborhoods. There have also been a large number of home sales at the southern edge of the city along Timberlake Road.
```{r}
#| label: fig-lburg_sales
#| fig-cap: "Map of home sales in Lynchburg from 2016 through 2022"

library(tidygeocoder)

# lburg_props <- read_excel("data/lburg_sold.xlsx", sheet = "Residential") %>% 
#   clean_names() %>% 
#   geocode(address = street_full_address,
#           method = 'geocodio',
#           lat = latitude,
#           long = longitude)

# write_rds(lburg_props, "data/lburg_props.rds")

library(zoo)

# Hex map has some of these properties that may be outside of city limits. Might need to explore this and filter those out. 

lburg_props <- read_rds("data/lburg_props.rds") %>% 
  mutate(date = as.yearmon(selling_date, "%Y %m")) %>% 
  mutate(date = as.Date(date)) %>% 
  mutate(year = year(date)) %>% 
  drop_na(latitude) 

lburg_props_pcovid <- lburg_props %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326)

library(mapview)

hex_grid <- st_make_grid(lburg_props_pcovid, n = c(25,25),
                         what = "polygons", square = FALSE)

hex_grid_sold <- st_sf(hex_grid) %>% 
  mutate(grid_id = 1:length(lengths(hex_grid)))

hex_grid_sold$n_pts <- lengths(st_intersects(hex_grid_sold, lburg_props_pcovid))

hex_count <- filter(hex_grid_sold, n_pts > 5)

mapview(hex_count, zcol = "n_pts")


```

## Rental market

Like the counties, rents in the city have been relatively flat from 2015 to 2020. From 2020 onward, there were bumps in the market asking rent that led to a high of \$1,106 in the second quarter of 2022. But average market asking rent has been in decline since then and sat at \$1,046 by the end of 2022, the lowest average rent over the course of seven years.

```{r}
#| label: fig-rent
#| fig-cap: "Quarterly average market asking rent in Lynchburg"

# Utilize CoStar properties search for Multi-Family. Filter by County for Lynchburg - VA Market. Then review Analytics and pull QUARTERLY Market Asking Rent Per Unit for entire region. Do not bring in Forecast and set History to All. Combine Bedford City and Bedford County.

# Download data for current rent and inflation-adjusted rent into a single csv - name [locality_abbv]_rent.csv --- for example amherst_rent.csv or lburg_rent.csv

# Merge all data into a single csv named - local_rent.csv

# Note the download date of data here: 1/23/2023

# Read in downloaded data for Lynchburg rent.

 lburg_rent <- read_csv("data/lburg_rent.csv") %>% 
  mutate(county = "Lynchburg city") # Create a field for the locality name.


library(zoo)
library(lubridate)

local_rent <- lburg_rent %>% 
  filter(!str_detect(period, "QTD")) %>% # Remove QTD values.
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) %>%  # Convert quarter dates to readable dates.
  select(county, period, current, inf) %>% 
  pivot_longer(3:4, # Pivot data to create tabular data.
               names_to = "value",
               values_to = "rent") %>% 
  mutate(value = case_when(
    value == "current" ~ "Current dollars",
    value == "inf" ~ "Inflation-adjusted dollars")) %>% 
  mutate(year = year(period)) %>% 
  filter(year >= 2015) |> 
  mutate(rent_label = dollar(rent)) |> 
  mutate(tooltip = c(paste0("Date: ", period, "\n Rent: ", rent_label)))
  

local_rent_plot <- ggplot(local_rent,
       aes(x = period,
           y = rent,
           linetype = value,
           color = value)) +
  geom_line(size = 0.5) + 
  geom_point_interactive(aes(
    data_id = tooltip,
    tooltip = tooltip)) +
  scale_linetype_manual(values = c("dotted", "solid")) +
  theme_hfv() +
  scale_color_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom") +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Quarterly average market asking rent in Lynchburg",
       caption = "**Source:** CoStar Group Inc.")
  

if (knitr::is_html_output()) {
  
  girafe(ggobj = local_rent_plot,
         height_svg = 4)
  
} else {
  
  local_rent_plot
  
}


```

```{r}
#| label: vacancy

# Utilize CoStar properties search for Multi-Family. Filter by County for Lynchburg - VA Market. Then review Analytics and pull QUARTERLY Vacancy Rate for entire region. Do not bring in Forecast and set History to All.

# Download data into csv - name [locality_abbv]_vacancy.csv --- for example amherst_vacancy.csv or lburg_vacancy.csv

# Note the download date of data here: 




```
It is important to note that CoStar data excludes a significant amount of single-family detached rentals, which has been growing in demand across the nation. In the city, there is an estimated 4,922 single-family homes that are being used as rentals. A majority of the detached SFR is located in Ward II, where the College Hill and Diamond Hill neighborhoods are located.

Rental prices for these are not easily discerned, as this type of housing is sometimes marketed on non-traditional channels, like Facebook groups or simply with a yard sign.

```{r}
#| label: parcel-data
#| eval: false


parcels <- st_read("https://mapviewer.lynchburgva.gov/ArcGIS/rest/services/OpenData/ODPDynamic/MapServer/41/query?outFields=*&where=1%3D1&f=geojson", quiet = TRUE) %>% 
  st_point_on_surface()

write_rds(parcels, "data/parcels.rds")

```


```{r}
#| label: fig-sfr
#| fig-cap: "Map of single-family detached rentals in Lynchburg"

# The following investigates the number of single-family detached homes that are being utilized as rentals. There is a growing concern over homeownership rates in the city. The current market incentivizes the rental of single-family homes over the sale.

# wards <- st_read("https://mapviewer.lynchburgva.gov/ArcGIS/rest/services/OpenData/ODPStatic/MapServer/62/query?outFields=*&where=1%3D1&f=geojson", quiet = TRUE)

# write_rds(wards, "data/wards.rds")

wards <- read_rds("data/wards.rds")

parcels <- read_rds("data/parcels.rds")

residential <- parcels %>%
  clean_names() %>% 
  select(parcel_id, loc_addr, loc_city, loc_zip, owner1, mail_addr, mail_city, mail_stat, mail_zip, pc_desc, yr_built, current_land, current_imp) %>% 
  filter(str_detect(pc_desc, "DETACHED SNG FAM")) %>% 
  st_join(left = FALSE, wards["Designation"])

library(stringdist)

residential$match_score <- stringsim(residential$loc_addr, residential$mail_addr)

# ggplot(residential,
#        aes(x = match_score)) +
#   geom_histogram(bins = 30) +
#   theme_minimal()
# 
# set.seed(1983)
# 
# fuzzy_matches <- residential %>%
#   filter(match_score >= 0.75 & match_score <= 0.9) %>%
#   st_drop_geometry() %>%
#   slice_sample(n = 100)

res_own <- residential %>% 
  mutate(address_num = str_extract(loc_addr, "[0-9]+"),
         owner_num = str_extract(mail_addr, "[0-9]+")) %>% 
  filter(current_imp > 0) %>% 
  mutate(ownership = case_when(
    match_score > 0.6 & address_num == owner_num ~ "Owner-occupied SFD",
    TRUE ~ "SFR")
  )

sfr <- res_own %>% 
  filter(ownership == "SFR")

hex_grid <- st_make_grid(sfr, n = c(25,25),
                         what = "polygons", square = FALSE)

hex_grid_sf <- st_sf(hex_grid) %>% 
  mutate(grid_id = 1:length(lengths(hex_grid)))

hex_grid_sf$n_pts <- lengths(st_intersects(hex_grid_sf, sfr))

hex_count <- filter(hex_grid_sf, n_pts > 4)

library(mapview)

mapview(hex_count,
        zcol = "n_pts",
        layer.name = "Estimate")

```

```{r}
#| label: fig-sfr-ward
#| fig-cap: "Single-family detached rentals by Lynchburg ward"

sfr_ward <- sfr %>% 
  group_by(Designation) %>% 
  summarise(count = n_distinct(parcel_id)) %>% 
  st_drop_geometry()

sfr_ward_plot <- ggplot(sfr_ward,
       aes(x = Designation,
           y = count,
           fill = Designation,
           data_id = count,
           tooltip = number_format(big.mark = ",")(count))) +
  geom_col(position = "dodge") +
  geom_col_interactive() +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Estimate of single-family detached rentals",
       caption = "**Source:** Lynchburg city parcel data") 

if (knitr::is_html_output()) {
  
  girafe(ggobj = sfr_ward_plot,
         height_svg = 4)
  
} else {
  
  sfr_ward_plot
  
}


```
As of early 2023, there were a total of 3,651 assisted rental units in the city. While assisted units can be supported by multiples types of federal subsidy, the bulk of these units (35 percent) are supported by the Low-Income Housing Tax Credit (LIHTC) program. They are followed by project-based Section 8 at 1,201 units. 

:::{.callout-note}

Add information about the status of public housing in Lynchburg.

:::
```{r}
#| label: fig-subsidy
#| fig-cap: "Federally-assisted rental homes in Lynchburg by subsidy"

nhpd_prop <- read_csv("data/nhpd_properties.csv") %>% 
  clean_names() %>% 
  filter(property_status != "Inactive") %>% 
  filter(grepl("lynchburg", city, ignore.case = TRUE))

nhpd_subsidies <- read_csv("data/nhpd_subsidies.csv") %>% 
  clean_names() %>% 
  filter(subsidy_status != "Inactive") %>% 
  filter(grepl("lynchburg", city, ignore.case = TRUE)) %>% 
  group_by(subsidy_name) %>% 
  summarise(units = sum(assisted_units))


subsidy_plot <- ggplot(nhpd_subsidies,
       aes(x = reorder(subsidy_name, -units),
           y = units,
           fill = subsidy_name,
           tooltip = label_number(big.mark = ",")(units),
           data_id = units)) +
  geom_col() +
  geom_col_interactive() +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Federally-assisted rental homes in Lynchburg by subsidy",
       caption = "**Source:** National Housing Preservation Database (NHPD).")


if (knitr::is_html_output()) {
  
  girafe(ggobj = subsidy_plot,
         height_svg = 4)
  
} else {
  
  subsidy_plot
}

```
As a major portion of Lynchburg's affordable housing stock, LIHTC is potentially at-risk due to expiring affordability commitments. By 2035, over 40 percent of active LIHTC units are set to exit their affordability commitment period. These properties would need to obtain an additional allocation of tax credits to maintain affordability. While nonprofit owner/operators will seek to maintain affordability, for-profit owners are more likely to convert their properties to market rate.
```{r}
#| label: fig-lihtc-exp
#| fig-cap: "LIHTC units by affordability commitment expiration in Lynchburg"


lihtc <- read_csv("data/nhpd_subsidies.csv") %>% 
  clean_names() %>% 
  filter(subsidy_status != "Inactive") %>% 
  filter(grepl("lynchburg", city, ignore.case = TRUE)) %>% 
  filter(subsidy_name == "LIHTC") %>% 
  mutate(end_date = mdy(end_date)) %>% 
  mutate(year = year(end_date)) %>% 
  group_by(year) %>% 
  summarise(units = sum(assisted_units)) %>% 
  mutate(pct = (units/sum(units)))

lihtc_plot <- ggplot(lihtc,
       aes(x = year,
           y = pct,
           fill = pct,
           data_id = pct,
           tooltip = percent_format()(pct)
           )) +
  geom_col() +
  geom_col_interactive() +
  theme_hfv() +
  labs(title = "LIHTC units by affordability commitment expiration date",
       subtitle = "Percent of total in Lynchburg",
       caption = "**Source:** National Housing Preservation Database (NHPD).") +
  scale_fill_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) +
  scale_y_continuous(labels = percent_format())



if (knitr::is_html_output()) {
  
  girafe(ggobj = lihtc_plot,
         height_svg = 4)
  
} else {
  
lihtc_plot

}

  

```

```{r}
#| label: gap-data
#| eval: false

tb_18c <- read_csv("data/Table18C_2012to2019.csv")

colnames(tb_18c)[16] <- "Cost"

renter <- tb_18c %>% 
  clean_names() %>% 
  filter(line_type == "Detail") %>% 
  select(county, fips, year, estimate, tenure, cost, household_income) %>% 
  group_by(county, fips, year, tenure, cost, household_income) %>% 
  summarise(estimate = sum(estimate))


tb_18_match <- renter %>% 
  filter(tenure == "Renter occupied") %>% 
  mutate(cost = case_when(
    cost == "greater than RHUD30 and less than or equal to RHUD50" ~ "31 to 50 percent AMI",
    cost == "greater than RHUD50 and less than or equal to RHUD80" ~ "51 to 80 percent AMI",
    cost == "greater than RHUD80" ~ "80 percent AMI or greater",
    cost == "less than or equal to RHUD30" ~ "30 percent AMI or below"
  )) %>% 
  mutate(household_income = case_when(
    household_income == "greater than 100% of HAMFI" ~ "80 percent AMI or greater",
    household_income == "greater than 80% of HAMFI but less than or equal to 100% of HAMFI" ~ "80 percent AMI or greater",
    household_income == "greater than 50% of HAMFI but less than or equal to 80% of HAMFI" ~ "51 to 80 percent AMI",
    household_income == "greater than 30% of HAMFI but less than or equal to 50% of HAMFI" ~ "31 to 50 percent AMI",
    household_income == "less than or equal to 30% of HAMFI" ~ "30 percent AMI or below"
  )) |>
  group_by(county, fips, year, cost, household_income) %>% 
  summarise(estimate = sum(estimate))


tb_18_match <- tb_18_match %>% 
  mutate(match = case_when(
    cost == household_income ~ "Affordable",
    cost == "30 percent AMI or below" & household_income == "31 to 50 percent AMI" ~ "Very affordable",
    cost == "30 percent AMI or below" & household_income == "51 to 80 percent AMI" ~ "Very affordable",
    cost == "30 percent AMI or below" & household_income == "80 percent AMI or greater" ~ "Very affordable",
        cost == "31 to 50 percent AMI" & household_income == "31 to 50 percent AMI" ~ "Affordable",
    cost == "31 to 50 percent AMI" & household_income == "51 to 80 percent AMI" ~ "Very affordable",
    cost == "31 to 50 percent AMI" & household_income == "80 percent AMI or greater" ~ "Very affordable",
        cost == "31 to 50 percent AMI" & household_income == "30 percent AMI or below" ~ "Unaffordable",
    cost == "51 to 80 percent AMI" & household_income == "30 percent AMI or below" ~ "Unaffordable",
        cost == "51 to 80 percent AMI" & household_income == "31 to 50 percent AMI" ~ "Unaffordable",
    cost == "51 to 80 percent AMI" & household_income == "80 percent AMI or greater" ~ "Very affordable",
    cost == "80 percent AMI or greater" & household_income == "30 percent AMI or below" ~ "Unaffordable",
    cost == "80 percent AMI or greater" & household_income == "31 to 50 percent AMI" ~ "Unaffordable",
    cost == "80 percent AMI or greater" & household_income == "51 to 80 percent AMI" ~ "Unaffordable"
  )) %>% 
  mutate(gapcode = case_when(
    match == "Unaffordable" ~ "Gap",
    TRUE ~ "Matches or less than income"
  ))

write_rds(tb_18_match, "data/tb_18_match.rds")

```

```{r}
#| label: fig-gap
#| fig-cap: "Rental housing gap by AMI in Lynchburg"

tb_18_match <- read_rds("data/tb_18_match.rds") %>% 
  filter(fips == lburg)


gap <- tb_18_match %>% 
  group_by(year, household_income, gapcode) %>% 
  summarise(estimate = sum(estimate)) %>% 
  mutate(estimate = case_when(
    gapcode == "Gap" ~ estimate * -1,
    gapcode == "Matches or less than income" ~ estimate
  )) %>% 
  filter(household_income != "80 percent AMI or greater") %>% 
  mutate(gapcode = case_when(
    gapcode == "Gap" ~ "Unaffordable",
    TRUE ~ "Affordable"
  )) %>% 
  mutate(year = as.character(year))

gap_plot <- ggplot(gap,
       aes(x = year, 
           y = estimate,
           fill = gapcode,
           data_id = estimate,
           tooltip = label_number(big.mark = ",")(estimate))) +
  geom_col() +
  geom_col_interactive() +
  facet_grid(~household_income) +
  scale_y_continuous(labels = label_comma()) +
  theme_hfv() +
  scale_fill_hfv() +
  labs(title = "Rental housing gap by AMI in Lynchburg",
       caption = "**Source:** U.S. Department of Housing and Urban Development, Comprehensive Housing Affordability Strategy, Table 18B") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, vjust = 0.5),
    axis.title = element_blank(),
    panel.background = element_blank(),
    panel.grid.major.y = element_line(color = "grey95",
                                      size = 0.05),
    axis.ticks = element_blank()
  )

if (knitr::is_html_output()) {
  
  girafe(ggobj = gap_plot,
         height_svg = 4)
  
} else {
  
  gap_plot
  
}


```

```{r}
#| label: hcv-data
#| eval: false

hcv <- st_read("https://services.arcgis.com/VTyQ9soqVukalItT/arcgis/rest/services/Housing_Choice_Vouchers_by_Tract/FeatureServer/0/query?f=json&where=(STATE%20IN%20('51'))&outFields=*", quiet = TRUE)

write_rds(hcv, "data/hcv.rds")

```

```{r}
#| label: fig-hcv-map
#| fig-cap: "Housing Choice Voucher utilization in Lynchburg"

hcv <- read_rds("data/hcv.rds") %>% 
  st_transform(crs = 4326) %>% 
  clean_names() %>% 
  filter(county == 680) 

library(mapview)

mapview(hcv, zcol = "hcv_pct_renter_occ_units")

# palette <- colorNumeric(palette = "viridis", domain = hcv$hcv_pct_renter_occ_units, na.color = "transparent")
# 
# leaflet(hcv) %>% 
#   addTiles() %>% 
#   addPolygons(fillColor = ~palette(hcv_pct_renter_occ_units),
#               fillOpacity = 1)


```


```{r}
#| label: rprops

# Utilize CoStar properties search for Multi-Family. Filter for City of Lynchburg.

# Download property-level data export, including lat/long.


```
## Affordability

In 2021, there was a total of 8,875 cost-burdened households in the city, 75 percent of which were renters. Even those the overall total of cost-burdened households in the city has declined by five percent, that decline has been mainly among homeowner households. 

The number of cost-burdened renter households has increased in the last decade from 5,380 to 6,663, a 24 percent increase.

```{r}
#| label: fig-local-cb
#| fig-cap: "Lynchburg cost burden by tenure"

cv_cb <- read_rds("data/b25106_data.rds") |> 
  filter(NAME == "Lynchburg city")

local_cb <- cv_cb |> 
  group_by(NAME, year, tenure, cb) |> 
  summarise(estimate = sum(estimate)) |> 
  filter(cb == "Cost-burdened") |> 
  mutate(year_label = as.character(year))


cv_cb_plot <- ggplot(local_cb,
       aes(x = year_label,
           y = estimate,
           fill = tenure,
           data_id = estimate,
           tooltip = label_number(big.mark = ",")(estimate))) +
  geom_col(position = "stack") +
  geom_col_interactive(position = "stack") +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_number(big.mark = ",")) + 
  labs(title = "Lynchburg cost-burdened households by tenure",
       caption = "**Source:** U.S. Census Bureau, ACS 5-Year Estimates, Table B25106.") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5))



if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_cb_plot,
         height_svg = 4)
} else {
  
  cv_cb_plot
}
  


```

```{r}
#| label: fig-cb-tenure
#| eval: false
#| fig-cap: "Lynchburg cost burden by tenure"

cb_tenure <- read_rds("data/cb_7.rds") %>% 
  subset(fips %in% lburg) %>% 
  filter(cb_group == "Cost-burdened") %>% 
  group_by(year, tenure, cb_group) %>% 
  summarise(estimate = sum(estimate)) %>% 
  group_by( tenure) |> 
  mutate(change = estimate - lag(estimate)) |> 
  drop_na() |> 
  mutate(run_diff = cumsum(change))

cb_diff <- ggplot(cb_tenure,
       aes(x = year, 
           y = run_diff, 
           fill = tenure,
           data_id = run_diff,
           tooltip = run_diff)) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge", size = 2) +
  theme_hfv() +
  scale_fill_hfv() +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Cumulative change in cost-burdened households by tenure",
       subtitle = "2012 to 2019",
       fill = "Tenure",
       caption = "**Source:** U.S Department of Housing and Urban Development,<br>Comprehensive Housing Affordability Strategy (CHAS), Table 7.") +
  theme(legend.position = "right") +
  add_zero_line(axis = "y")


if (knitr::is_html_output()) {
  
  girafe(ggobj = cb_diff,
         height_svg = 4) 
  
} else {
  
  cb_diff
  
  }
  




```

```{r}
#| label: fig-cb-race
#| fig-cap: "Lynchburg cost burden by race and ethnicity"

cb_9 <- read_rds("data/cb_9.rds")

cv_cb9 <- cb_9 %>% 
  subset(fips %in% lburg) %>% 
  group_by(year, race, cb_group) %>% 
  summarise(estimate = sum(estimate)) %>% 
  group_by(year, race) %>% 
  mutate(pct = estimate/sum(estimate)) %>% 
  filter(cb_group == "Cost-burdened")

cv_cb9_plot <- ggplot(cv_cb9,
       aes(x = year, 
           y = pct,
           fill = str_wrap(race, width = 10),
           data_id = pct,
           tooltip = percent_format()(pct))) +
  geom_col(position = "stack") +
  geom_col_interactive(size = 2) +
  facet_wrap(~race, nrow = 1, labeller = label_wrap_gen(8)) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_hfv() +
  theme_hfv() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        strip.text.x = element_text(size = 8),
        strip.text.y = element_blank(),
        axis.text.y = element_blank()) +
  labs(title = "Cost burden by race and ethnicity",
       subtitle = "Percent of cost-burdened households",
       caption = "**Source:** U.S. Department of Housing and Urban Development, Comprehensive Housing Affordability Strategy (CHAS), Table 9.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = cv_cb9_plot,
         height_svg = 4)
  
} else {
  
  cv_cb9_plot
  
}


```



